---
pdf_document: default
title: "A549 +/- GSK591 2days (PRMT5i) Fractionation RNA-Seq with Dm6 spike-in"
author: "Maxim Maron, David Shechter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: null
  df_print: paged
  pdf_document: default
html_notebook: default
---

Data is loaded from pre-computed Salmon quant outputs

*This analysis requires STAR mapping to a joint hg38-dm6 genome and a pre-built gtf with "dm6" in the Chromsome column for any control spike-in gene. These files are available at [www.github.com/shechterlab](www.github.com/shechterlab)*

```{r "setup", include=FALSE}

# Set the WD here
working_dir <- 'C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 6/'
knitr::opts_knit$set(root.dir = normalizePath(working_dir))
setwd(working_dir)
getwd()

#load librarys
library(tidyverse)
library(DESeq2)
library(tximport)
library(apeglm)
library(EnhancedVolcano)
library(RColorBrewer)
library(pheatmap)
library(knitr)
library(kableExtra)
library(ggplot2)

# General function to load data
load_data <- function(file_path, delimiter = ",", header = TRUE) {
  read.delim(file_path, sep = delimiter, header = header)
}

#For DESeq2 - count threshold
count_cutoff = 20

#For QC output only
pval_threshold = 0.01
log2FC_threshold = 1 

#this tx2gene and combined gtf file was independently generated, necessary for spike-in analysis
dm6.hg38.tx2gene_file = 'dm6.hg38.tx2gene.csv'
gtf_gene_annotation_table = 'dm6.hg38.gtf_gene_annotation_table.txt'


#initialize this list for filling during the function
summary_stats_list <- list()

#create output dir
if (!file.exists(paste0(working_dir, "DESeq2_output/"))) {
  dir.create(working_dir, "DESeq2_output") }

```

```{r "Define Functions", include=FALSE}

# DESeq2 Analysis Function
#This function performs DESeq2 analysis, generates plots, saves the results, and provides summary statistics.

#used if not knitting
setwd(working_dir)

# General function to run DESeq2 analysis, with features as indicated
run_deseq2_analysis <- function(txi, col_data, condition_ref, features, output_prefix, use_spikeIn, pvalue_threshold = pval_threshold) {
  
  dds <- DESeqDataSetFromTximport(txi, colData = col_data, design = ~ condition)
  
  if (use_spikeIn) {
  
      # Check if 'Chromosome' column exists in features
    if(!"Chromosome" %in% colnames(features)) {
      stop("The features data frame does not contain a 'Chromosome' column.")
    }
    
    #Define ctrlGenes from all gene annotated with dm6 in the Chromosome column
    
    
    ctrlGenes <- rownames(dds) %in% unique(features %>% filter(grepl("dm6", Chromosome)) %>% pull(gene_id))
    if (length(ctrlGenes) == 0) {
      stop("No control genes found with 'dm6' in the Chromosome column.")
    }
    dds_ctrl <- dds[ctrlGenes,]
    dds_ctrl <- dds_ctrl[rowSums(counts(dds_ctrl) >= count_cutoff),]
    
    nm <- assays(dds_ctrl)[["avgTxLength"]]
    counts <- counts(dds_ctrl)
    sf <- counts / nm
    sf <- sf[!duplicated(sf),]
    sf <- estimateSizeFactorsForMatrix(sf)
    
    #Output scale factor for each run
    print(paste0("Scale factors for", output_prefix, "_vs_", condition_ref))
    print(sf)
    
    #Define tested genes as those without dm6 in Chromosome column
    testGenes <- rownames(dds) %in% unique(features %>% filter(!grepl("dm6", Chromosome)) %>% pull(gene_id))
    
    if (length(testGenes) == 0) {
      stop("No test genes found without 'dm6' in the Chromosome column.")
    }
    dds <- dds[testGenes,]
    #dds <- dds_ctrl[rowSums(counts(dds) >= count_cutoff),]
    keep <- rowSums(counts(dds)) >= count_cutoff
    dds <- dds[keep,]
    sizeFactors(dds) <- sf
  } 
  
  else {
    # Normalize using avgTxLength only, not with ctrlGenes spik In
    nm <- assays(dds)[["avgTxLength"]]
    counts <- counts(dds)
    sf <- counts / nm
    sf <- estimateSizeFactorsForMatrix(sf)
    sizeFactors(dds) <- sf
    
    # Filter out dm6 genes
    
    if(!"Chromosome" %in% colnames(features)) {
      stop("The features data frame does not contain a 'Chromosome' column.")
    }
    
    #Define tested genes as those without dm6 in Chromosome column
    testGenes <- rownames(dds) %in% unique(features %>% filter(!grepl("dm6", Chromosome)) %>% pull(gene_id))
    
    if (length(testGenes) == 0) {
      stop("No test genes found without 'dm6' in the Chromosome column.")
    }
    
    dds <- dds[testGenes,]
    keep <- rowSums(counts(dds)) >= count_cutoff
    dds <- dds[keep,]
  }
  
  dds$condition <- relevel(dds$condition, ref = condition_ref)
  #Regular wald test
  dds <- DESeq(dds, test = "Wald")
  
  #uncomment above to reset
  #Test with likelihood ratio test hypothesis
  #dds <- DESeq(dds, test = "LRT", reduced = ~1)
  
  vsd <- vst(dds, blind = FALSE)
  
  # setup for saving output files
  
  # Create output directory
  output_dir <- paste0("Output")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir)
  }
  
  # Setup output files
  output_prefix <- file.path(output_dir, output_prefix)
  
  # Heatmap - uncomment if you want it, generally not useful
    #pheatmap(assay(vsd),
  #         cluster_rows = FALSE,
  #         show_rownames = FALSE,
  #         cluster_cols = FALSE,
  #         annotation_col = as.data.frame(colData(dds)[,"condition"], row.names=rownames(colData(dds))))
  
  sampleDists <- dist(t(assay(vsd)))
  sampleDistMatrix <- as.matrix(sampleDists)
  rownames(sampleDistMatrix) <- paste(vsd$condition, sep="-")
  colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
  
  # Sample distance heatmap
  pheatmap(sampleDistMatrix,
           clustering_distance_rows = sampleDists,
           clustering_distance_cols = sampleDists,
           col = colors)
  
  # PCA plot
  plotPCA(vsd, intgroup = c("condition"))
  
  for (i in 2:length(unique(resultsNames(dds)))){
    name <- unique(resultsNames(dds))[i]
    res <- lfcShrink(dds, coef = name, type = "apeglm")
    res$gene_id <- rownames(res)
    res_df <- as.data.frame(res)
    res_df <- merge(res_df, features, by.x = "gene_id", by.y = "gene_id")
    res_df <- res_df[order(res_df$padj),]
    
    #COMMENT OUT BELOW TWO LINES IF USING BH CORRECTION  
    # Apply Bonferroni correction manually
    #   res_df$padj <- p.adjust(res_df$pvalue, method = "bonferroni")
    #   res_df <- res_df[order(res_df$padj),]
    
    if(use_spikeIn) {
      write.csv(res_df, file = paste0(working_dir, output_prefix, "_vs_", condition_ref, "_SpikeInNormalized_DESEQ2.csv"), row.names = FALSE) }
    else {
    write.csv(res_df, file = paste0(working_dir, output_prefix, "_vs_", condition_ref, "_noSpikeIn_DESEQ2.csv"), row.names = FALSE) 
    }
    
    # MA plot
    plotMA(res)
    
    res_df <- res_df[!is.na(res_df$padj),]
    
    # Volcano plot
    volcano <- EnhancedVolcano(res_df,
                          lab = res_df$GeneSymbol,
                          x = 'log2FoldChange',
                          y = 'padj',
                          title = name,
                          pCutoff = pvalue_threshold,
                          FCcutoff = log2FC_threshold,
                          pointSize = 2.0,
                          labSize = 5.0,
                          col = c('#d4d4d4', '#d4d4d4', '#fdbb84', '#e34a33'),
                          colAlpha = 0.5)
    
    print(volcano)
    
    if(use_spikeIn) {
    ggsave(filename = paste0(working_dir, output_prefix, "_vs_", condition_ref, "_SpikeInNormalized_VolcanoPlot.pdf"), plot = volcano)
    }
    else {
      ggsave(filename = paste0(working_dir, output_prefix, "_vs_", condition_ref, "_noSpikeIn_VolcanoPlot.pdf"), plot = volcano)
    }
    # P-value histogram
    pval_hist <- ggplot(res_df, aes(x = pvalue)) + 
      geom_histogram(binwidth = 0.05, fill = "darkgray", color = "black") + 
      theme_minimal() +
      labs(title = paste0("Raw P-value Histogram: ", name), x = " P-value", y = "Frequency")
    print(pval_hist)
    
      # P-value histogram
    pval_hist <- ggplot(res_df, aes(x = padj)) + 
      geom_histogram(binwidth = 0.05, fill = "darkgray", color = "black") + 
      theme_minimal() +
      labs(title = paste0("BH Adjusted P-value Histogram: ", name), x = " P-value", y = "Frequency")
    print(pval_hist)
    
     # Summary statistics
    summary_stats <- res_df %>% 
      summarise(
        total_genes = n(),
        significant_genes = sum(padj < pvalue_threshold, na.rm = TRUE),
        upregulated_genes = sum(log2FoldChange > 0 & padj < pvalue_threshold, na.rm = TRUE),
        downregulated_genes = sum(log2FoldChange < 0 & padj < pvalue_threshold, na.rm = TRUE)
      )
    
    print(knitr::kable(summary_stats, caption = paste0("Summary Statistics for ", name)) %>% 
            kableExtra::kable_styling(bootstrap_options = "striped", full_width = F))
    
   }
}

# Function to call DESeq2

call_DESeq2_analysis <- function(test_sample_name, reference_sample_name, prefix, use_spikeIn) {

    # Filter files and samples for both test and reference conditions
  
  testfiles <- files[grepl(test_sample_name, names(files))]
  referencefiles <- files[grepl(reference_sample_name, names(files))]
  
  samples_to_test <- samples[grepl(test_sample_name, samples$condition) | grepl(reference_sample_name, samples$condition),]
  
  # Combine test and reference files
  combined_files <- c(referencefiles, testfiles)
  
  # Debugging output
  #print(paste("Number of combined files:", length(combined_files)))
  #print(paste("Number of samples to test:", nrow(samples_to_test)))
  #print("Combined files:")
  #print(combined_files)
  #print("Samples to test:")
  #print(samples_to_test)
  
  # Ensure the lengths of combined_files and samples_to_test match
  if(length(combined_files) != nrow(samples_to_test)) {
    stop("The number of combined files does not match the number of samples to test.")
  }
  
  # Assign names to combined_files
  names(combined_files) <- samples_to_test$sample
  
  # Debugging output to verify names assignment
 # print("Combined files with assigned names:")
#  print(combined_files)
  
  txi <- tximport(combined_files, type = "salmon", tx2gene = tx2gene)
  
  run_deseq2_analysis(txi, samples_to_test, reference_sample_name, features, prefix, use_spikeIn = use_spikeIn)
}



```


```{r "Load Data", include=FALSE}
#This section loads the necessary data files.
# Load the necessary files
tx2gene <- load_data(dm6.hg38.tx2gene_file)

#Load in sample metadata
samples <- read.csv(paste0(working_dir, 'samplesheet_DESeq2.csv'), header = T)

#path to gtf_gene_annotation_table.txt
features <- read.table(gtf_gene_annotation_table, header =T, fill = T)

# Create file paths for quant.sf files
files <- file.path(working_dir, "salmon_quant", samples$sample, "quant.sf")
names(files) <- samples$sample
```

***

## HEADING/TITLE FOR THIS ANALYSIS HERE ##

## GSK591 vs DMSO in Chromatin with SpikeIn Normalization ##

```{r, "GSK591 vs DMSO in chromatin", warning=FALSE}

test_sample_name <- "GSK591_chromatin"
reference_sample_name <- "DMSO_chromatin"
prefix <- test_sample_name
use_spikeIn = TRUE

call_DESeq2_analysis(test_sample_name, reference_sample_name, prefix, use_spikeIn)

```

***


## GSK591 vs DMSO in Nucleoplasm with SpikeIn Normalization ##

```{r, "GSK591 vs DMSO in Nucleoplasm", warning=FALSE}

test_sample_name <- "GSK591_nucleoplasm"
reference_sample_name <- "DMSO_nucleoplasm"
prefix <- test_sample_name
use_spikeIn = TRUE

call_DESeq2_analysis(test_sample_name, reference_sample_name, prefix, use_spikeIn)

```



## GSK591 vs DMSO in Cytoplasm with SpikeIn Normalization ##

```{r, "GSK591 vs DMSO in Cytoplasm", warning=FALSE}

test_sample_name <- "GSK591_cytoplasm"
reference_sample_name <- "DMSO_cytoplasm"
prefix <- test_sample_name
use_spikeIn = TRUE

call_DESeq2_analysis(test_sample_name, reference_sample_name, prefix, use_spikeIn)

```


## DMSO Chromatin vs DMSO Cytoplasm - with SpikeIn Normalization ##

```{r, "DMSO Chromatin vs DMSO cytoplasm with SpikeIn", warning=FALSE}

test_sample_name <- "DMSO_chromatin"
reference_sample_name <- "DMSO_cytoplasm"
prefix <- test_sample_name
use_spikeIn = TRUE

call_DESeq2_analysis(test_sample_name, reference_sample_name, prefix, use_spikeIn)

```

## GSK591 Chromatin vs GSK591 Cytoplasm - with SpikeIn Normalization ##

```{r, "GSK591 Chromatin vs GSK591 cytoplasm with SpikeIn", warning=FALSE}

test_sample_name <- "GSK591_chromatin"
reference_sample_name <- "GSK591_cytoplasm"
prefix <- test_sample_name
use_spikeIn = TRUE

call_DESeq2_analysis(test_sample_name, reference_sample_name, prefix, use_spikeIn)

```

## DMSO chromatin vs DMSO in nucleoplasm - with SpikeIn Normalization ##

```{r, "DMSO chromatin  vs DMSO in nucleoplasm with SpikeIn", warning=FALSE}

test_sample_name <- "DMSO_chromatin"
reference_sample_name <- "DMSO_nucleoplasm"
prefix <- test_sample_name
use_spikeIn = TRUE

call_DESeq2_analysis(test_sample_name, reference_sample_name, prefix, use_spikeIn)

```


## GSK591 chromatin vs GSK591 in nucleoplasm - with SpikeIn Normalization ##

```{r, "GSK591 chromatin  vs GSK591 in nucleoplasm with SpikeIn", warning=FALSE}

test_sample_name <- "GSK591_chromatin"
reference_sample_name <- "GSK591_nucleoplasm"
prefix <- test_sample_name
use_spikeIn = TRUE

call_DESeq2_analysis(test_sample_name, reference_sample_name, prefix, use_spikeIn)

```


