---
title: "GRIPPs and PRMT5-GRIPPs Definition and Comparative Analysis"
author: "David Shechter"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document: default
  pdf_document: default
---

# GRIPPs Definition and Analysis

## Overview

## Criteria for Defining GRIPPs

This script defines and analyzes **Genomically-Retained Incompletely Processed Polyadenylated Transcripts (GRIPPs)** using RNA-seq data from A549 cells fractionated into chromatin and cytoplasmic compartments under DMSO control conditions. GRIPPs are identified as transcripts that exhibit a significant enrichment in the chromatin fraction compared to the cytoplasm, normalized using spike-in controls.

```{r setup, include=FALSE}
library(tidyverse)
library(ggrepel)
library(ggplot2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(biomaRt)
library(GenomicRanges)
library(AnnotationDbi)
library(rtracklayer)
library(data.table)
library(eulerr)
library(ggpubr)
library(stringr)

# Set working directory
working_dir <- "C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 6/GRIPPs"
setwd(working_dir)
output_dir <- file.path(working_dir, "output")
dir.create(output_dir, showWarnings = FALSE)

# Input files
DMSO_chromatin_vs_cytoplasm_file <- "../Output/DMSO_chromatin_vs_DMSO_cytoplasm_SpikeInNormalized_DESEQ2.csv"
prmt5_chromatin_vs_cytoplasm_file <- "../Output/GSK591_chromatin_vs_GSK591_cytoplasm_SpikeInNormalized_DESEQ2.csv"
gencode_gtf <- "../data/gencode.v48.annotation.gtf"
chromHMM_bed <- "../data/ENCFF343YVN_A549-hg38_chromHMM-18-EtOH.bed"

# Cutoffs
cutoff <- list(log2FC = 2, padj = 0.01)
```

## Define GRIPPs and PRMT5-GRIPPs

```{r define-gripps}
dmso_res <- read.csv(DMSO_chromatin_vs_cytoplasm_file)
gsk_res <- read.csv(prmt5_chromatin_vs_cytoplasm_file)

# Strip Ensembl version suffixes if present
dmso_res$gene_id <- gsub("\\..*", "", dmso_res$gene_id)
gsk_res$gene_id <- gsub("\\..*", "", gsk_res$gene_id)

gripps <- dplyr::filter(dmso_res, log2FoldChange > cutoff$log2FC, padj < cutoff$padj) %>%
  dplyr::distinct(gene_id, .keep_all = TRUE)

prmt5_gripps <- dplyr::filter(gsk_res, log2FoldChange > cutoff$log2FC, padj < cutoff$padj) %>%
  dplyr::distinct(gene_id, .keep_all = TRUE)

# Diagnostic: Check overlap with GTF
gtf <- import(gencode_gtf)
gtf$gene_id <- gsub("\\..*", "", gtf$gene_id)
gtf_gene_ids <- unique(gtf$gene_id)
cat("GRIPPs gene IDs:", length(unique(gripps$gene_id)), "
")
cat("PRMT5-GRIPPs gene IDs:", length(unique(prmt5_gripps$gene_id)), "
")
cat("GRIPPs matched in GTF:", length(intersect(gripps$gene_id, gtf_gene_ids)), "
")
cat("PRMT5-GRIPPs matched in GTF:", length(intersect(prmt5_gripps$gene_id, gtf_gene_ids)), "
")

write.csv(gripps, file.path(output_dir, "GRIPPs.csv"), row.names = FALSE)
write.csv(prmt5_gripps, file.path(output_dir, "PRMT5-GRIPPs.csv"), row.names = FALSE)
```

## Overlap of GRIPPs and PRMT5-GRIPPs

```{r venn-diagram}
venn_counts <- list(GRIPPs = gripps$gene_id, `PRMT5-GRIPPs` = prmt5_gripps$gene_id)
eul <- euler(venn_counts)
venn_plot <- plot(eul, 
                  fills = list(fill = c("gray", "#2D8942"), alpha = 0.8), 
                  quantities = TRUE, 
                  labels = list(font = 4))


print(venn_plot)

ggsave(file.path(output_dir, "venn_GRIPPs_PRMT5-GRIPPs.pdf"), plot = venn_plot, device = "pdf", width = 5, height = 5)

```

## Generate BED Files for Each Group Separately

```{r generate-intron-bed}
#gtf <- import(gencode_gtf)
#gtf imported and fixed above

exons <- gtf[gtf$type == "exon"]

# Calculate introns once for all genes with exons
gtf$gene_id <- gsub("\\..*", "", gtf$gene_id)
exons$gene_id <- gsub("\\..*", "", exons$gene_id)

all_gene_ids <- unique(c(gripps$gene_id, prmt5_gripps$gene_id))
exons_by_gene <- split(exons[exons$gene_id %in% all_gene_ids], exons$gene_id)

introns_list <- lapply(names(exons_by_gene), function(gid) {
  gr <- exons_by_gene[[gid]]
  gr <- sort(gr)
  if (length(gr) < 2) return(NULL)
  gaps_gr <- gaps(gr)
  gaps_gr <- gaps_gr[strand(gaps_gr) == strand(gr)[1] & seqnames(gaps_gr) == seqnames(gr)[1]]
  gaps_gr <- gaps_gr[start(gaps_gr) >= min(start(gr)) & end(gaps_gr) <= max(end(gr))]
  if (length(gaps_gr) == 0) return(NULL)
  mcols(gaps_gr)$gene_id <- gid
  return(gaps_gr)
})

introns_list <- introns_list[!sapply(introns_list, is.null)]
introns_combined <- do.call(c, introns_list)
export.bed(introns_combined, file.path(output_dir, "All_introns.bed"))

# Subset for each group
gripps_introns <- introns_combined[mcols(introns_combined)$gene_id %in% gripps$gene_id]
prmt5_introns <- introns_combined[mcols(introns_combined)$gene_id %in% prmt5_gripps$gene_id]

export.bed(gripps_introns, file.path(output_dir, "GRIPPs_introns.bed"))
export.bed(prmt5_introns, file.path(output_dir, "PRMT5-GRIPPs_introns.bed"))
```

```{r generate-promoter-bed}

tx <- gtf[gtf$type == "transcript"]
get_promoters_bed <- function(gene_ids, label) {
  tx_subset <- tx[tx$gene_id %in% gene_ids]
  if (length(tx_subset) == 0) return(NULL)
  tx_promoters <- promoters(tx_subset, upstream = 500, downstream = 500)
  tx_promoters <- tx_promoters[!duplicated(tx_promoters$gene_id)]
  if (length(tx_promoters) == 0) return(NULL)
  promoter_df <- as.data.frame(tx_promoters)[,c("seqnames", "start", "end")]
  promoter_df$gene_id <- tx_promoters$gene_id
  promoter_df$gene_symbol <- tx_promoters$gene_name
  promoter_df$strand <- as.character(strand(tx_promoters))
  write.table(promoter_df, file.path(output_dir, paste0(label, "_promoters.bed")),
              sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}

get_promoters_bed(gripps$gene_id, "GRIPPs")
get_promoters_bed(prmt5_gripps$gene_id, "PRMT5-GRIPPs")


```

## ChromHMM Enrichment of Promoters

```{r chromHMM-promoters}

plot_chromHMM <- function(label) {
  bed_file <- file.path(output_dir, paste0(label, "_promoters.bed"))
  if (!file.exists(bed_file)) return(NULL)
  promoters_bed <- read.table(bed_file, header = FALSE)
  if (nrow(promoters_bed) == 0) return(NULL)
  colnames(promoters_bed) <- c("chrom", "start", "end", "gene_id", "gene_symbol", "strand")

  chromHMM <- fread(chromHMM_bed)
  colnames(chromHMM) <- c("chrom", "start", "end", "state", "score", "strand", "thickStart", "thickEnd", "rgb")

  promoters_gr <- GRanges(seqnames = promoters_bed$chrom,
                          ranges = IRanges(promoters_bed$start, promoters_bed$end))
  states_gr <- GRanges(seqnames = chromHMM$chrom,
                       ranges = IRanges(chromHMM$start, chromHMM$end),
                       state = chromHMM$state)

  hits <- findOverlaps(promoters_gr, states_gr)
  overlap_states <- mcols(states_gr)$state[subjectHits(hits)]
  overlap_df <- data.frame(state = overlap_states) %>% dplyr::count(state)

  state_totals <- as.data.frame(mcols(states_gr)) %>% dplyr::count(state)
  overlap_df <- dplyr::left_join(overlap_df, state_totals, by = "state") %>%
    dplyr::mutate(percent = 100 * n.x / n.y)

  ggplot(overlap_df, aes(x = reorder(state, -percent), y = percent)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    labs(title = paste("Promoter overlap with ChromHMM states:", label),
         x = "ChromHMM State", y = "% Overlap")
}

plot_chromHMM("GRIPPs")
plot_chromHMM("PRMT5-GRIPPs")

```

## GO Enrichment for GRIPPs and PRMT5-GRIPPs

```{r go-enrichment}
# CompareCluster to combine both groups into one enrichment plot
combined_gene_list <- list(
  GRIPPs = gripps$gene_id,
  `PRMT5-GRIPPs` = prmt5_gripps$gene_id
)

go_compare <- compareCluster(
  geneCluster = combined_gene_list,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db,
  keyType = "ENSEMBL",
  ont = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE
)

compare_plot <- enrichplot::dotplot(go_compare, showCategory = 10) +
  ggtitle("GO Biological Process Enrichment: GRIPPs vs PRMT5-GRIPPs")

print(compare_plot)
ggsave(file.path(output_dir, "GO_compare_GRIPPs_PRMT5-GRIPPs.pdf"), plot = compare_plot, width = 8, height = 6)
```
