---
title: "GRIPPs and PRMT5-GRIPPs Definition and Comparative Analysis"
author: "David Shechter"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document: default
  pdf_document: default
---

# GRIPPs Definition and Analysis

## Overview

## Criteria for Defining GRIPPs

This script defines and analyzes **Genomically-Retained Incompletely Processed Polyadenylated Transcripts (GRIPPs)** using RNA-seq data from A549 cells fractionated into chromatin and cytoplasmic compartments under DMSO control conditions. GRIPPs are identified as transcripts that exhibit a significant enrichment in the chromatin fraction compared to the cytoplasm, normalized using spike-in controls.

```{r setup, include=FALSE}
library(tidyverse)
library(ggrepel)
library(ggplot2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(biomaRt)
library(GenomicRanges)
library(AnnotationDbi)
library(rtracklayer)
library(data.table)
library(eulerr)
library(ggpubr)
library(stringr)

# Set working directory
working_dir <- "C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 6/GRIPPs"
setwd(working_dir)
output_dir <- file.path(working_dir, "output")
dir.create(output_dir, showWarnings = FALSE)

# Input files
DMSO_chromatin_vs_cytoplasm_file <- "../Output/DMSO_chromatin_vs_DMSO_cytoplasm_SpikeInNormalized_DESEQ2.csv"
prmt5_chromatin_vs_cytoplasm_file <- "../Output/GSK591_chromatin_vs_GSK591_cytoplasm_SpikeInNormalized_DESEQ2.csv"
gencode_gtf <- "../data/gencode.v48.annotation.gtf"
chromHMM_bed <- "../data/ENCFF343YVN_A549-hg38_chromHMM-18-EtOH.bed"

# Cutoffs
cutoff <- list(log2FC = 2, padj = 0.01)
```

## Define GRIPPs and PRMT5-GRIPPs

```{r define-gripps}
dmso_res <- read.csv(DMSO_chromatin_vs_cytoplasm_file)
gsk_res <- read.csv(prmt5_chromatin_vs_cytoplasm_file)

# Strip Ensembl version suffixes if present
dmso_res$gene_id <- gsub("\\..*", "", dmso_res$gene_id)
gsk_res$gene_id <- gsub("\\..*", "", gsk_res$gene_id)

gripps <- dplyr::filter(dmso_res, log2FoldChange > cutoff$log2FC, padj < cutoff$padj) %>%
  dplyr::distinct(gene_id, .keep_all = TRUE)

prmt5_gripps <- dplyr::filter(gsk_res, log2FoldChange > cutoff$log2FC, padj < cutoff$padj) %>%
  dplyr::distinct(gene_id, .keep_all = TRUE)

# Diagnostic: Check overlap with GTF
gtf <- import(gencode_gtf)
gtf$gene_id <- gsub("\\..*", "", gtf$gene_id)
gtf_gene_ids <- unique(gtf$gene_id)
cat("GRIPPs gene IDs:", length(unique(gripps$gene_id)), "
")
cat("PRMT5-GRIPPs gene IDs:", length(unique(prmt5_gripps$gene_id)), "
")
cat("GRIPPs matched in GTF:", length(intersect(gripps$gene_id, gtf_gene_ids)), "
")
cat("PRMT5-GRIPPs matched in GTF:", length(intersect(prmt5_gripps$gene_id, gtf_gene_ids)), "
")

write.csv(gripps, file.path(output_dir, "GRIPPs.csv"), row.names = FALSE)
write.csv(prmt5_gripps, file.path(output_dir, "PRMT5-GRIPPs.csv"), row.names = FALSE)
```

## Overlap of GRIPPs and PRMT5-GRIPPs

```{r venn-diagram}
venn_counts <- list(GRIPPs = gripps$gene_id, `PRMT5-GRIPPs` = prmt5_gripps$gene_id)
eul <- euler(venn_counts)

venn_plot <- plot(eul, 
                  fills = list(fill = c("gray", "#2D8942"), alpha = 0.8), 
                  quantities = TRUE, 
                  labels = list(font = 4))


print(venn_plot)

ggsave(file.path(output_dir, "venn_GRIPPs_PRMT5-GRIPPs.pdf"), plot = venn_plot, device = "pdf", width = 5, height = 5)

```

## Generate BED Files for Each Group Separately

```{r generate-intron-bed}
#gtf <- import(gencode_gtf)
#gtf imported and fixed above

exons <- gtf[gtf$type == "exon"]

# Calculate introns once for all genes with exons
gtf$gene_id <- gsub("\\..*", "", gtf$gene_id)
exons$gene_id <- gsub("\\..*", "", exons$gene_id)

all_gene_ids <- unique(c(gripps$gene_id, prmt5_gripps$gene_id))
exons_by_gene <- split(exons[exons$gene_id %in% all_gene_ids], exons$gene_id)

introns_list <- lapply(names(exons_by_gene), function(gid) {
  gr <- exons_by_gene[[gid]]
  gr <- sort(gr)
  if (length(gr) < 2) return(NULL)
  gaps_gr <- gaps(gr)
  gaps_gr <- gaps_gr[strand(gaps_gr) == strand(gr)[1] & seqnames(gaps_gr) == seqnames(gr)[1]]
  gaps_gr <- gaps_gr[start(gaps_gr) >= min(start(gr)) & end(gaps_gr) <= max(end(gr))]
  if (length(gaps_gr) == 0) return(NULL)
  mcols(gaps_gr)$gene_id <- gid
  return(gaps_gr)
})

introns_list <- introns_list[!sapply(introns_list, is.null)]
introns_combined <- do.call(c, introns_list)
export.bed(introns_combined, file.path(output_dir, "All_introns.bed"))

# Subset for each group
gripps_introns <- introns_combined[mcols(introns_combined)$gene_id %in% gripps$gene_id]
prmt5_introns <- introns_combined[mcols(introns_combined)$gene_id %in% prmt5_gripps$gene_id]

export.bed(gripps_introns, file.path(output_dir, "GRIPPs_introns.bed"))
export.bed(prmt5_introns, file.path(output_dir, "PRMT5-GRIPPs_introns.bed"))
```

```{r generate-promoter-bed}

tx <- gtf[gtf$type == "transcript"]
get_promoters_bed <- function(gene_ids, label) {
  tx_subset <- tx[tx$gene_id %in% gene_ids]
  if (length(tx_subset) == 0) return(NULL)
  tx_promoters <- promoters(tx_subset, upstream = 500, downstream = 500)
  tx_promoters <- tx_promoters[!duplicated(tx_promoters$gene_id)]
  if (length(tx_promoters) == 0) return(NULL)
  promoter_df <- as.data.frame(tx_promoters)[,c("seqnames", "start", "end")]
  promoter_df$gene_id <- tx_promoters$gene_id
  promoter_df$gene_symbol <- tx_promoters$gene_name
  promoter_df$strand <- as.character(strand(tx_promoters))
  write.table(promoter_df, file.path(output_dir, paste0(label, "_promoters.bed")),
              sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}

get_promoters_bed(gripps$gene_id, "GRIPPs")
get_promoters_bed(prmt5_gripps$gene_id, "PRMT5-GRIPPs")


```


## GO Enrichment and Gene Biotype Summary

```{r go-enrichment}
# Convert gene IDs to ENTREZ for clusterProfiler
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(biomaRt)

# ID conversion
gene_conversion <- bitr(
  unique(c(gripps$gene_id, prmt5_gripps$gene_id)),
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# Add group label
gene_conversion$group <- ifelse(
  gene_conversion$ENSEMBL %in% gripps$gene_id,
  "GRIPPs", "PRMT5-GRIPPs"
)

# CompareCluster for BP, MF, CC
cc_bp <- compareCluster(
  ENTREZID ~ group,
  data = gene_conversion,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE
)

cc_mf <- compareCluster(
  ENTREZID ~ group,
  data = gene_conversion,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db,
  ont = "MF",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE
)

cc_cc <- compareCluster(
  ENTREZID ~ group,
  data = gene_conversion,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db,
  ont = "CC",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE
)

# KEGG pathway enrichment
kegg_cc <- compareCluster(
  ENTREZID ~ group,
  data = gene_conversion,
  fun = "enrichKEGG",
  organism = "hsa",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05
)

# Plot and save
plots <- list(
  BP = cc_bp,
  MF = cc_mf,
  CC = cc_cc,
  KEGG = kegg_cc
)

for (ont in names(plots)) {
  p <- enrichplot::dotplot(plots[[ont]], showCategory = 8) +
    ggtitle(paste("Enrichment:", ont, "- GRIPPs vs PRMT5-GRIPPs"))
  print(p)
  ggsave(file.path(output_dir, paste0("GO_compare_", ont, "_GRIPPs_PRMT5-GRIPPs.pdf")),
         plot = p, width = 8, height = 6)
}
```

```{r gene-biotype-summary}
# Get gene biotypes
mart <- useMart("ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "https://www.ensembl.org")

biotypes <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
  filters = "ensembl_gene_id",
  values = unique(c(gripps$gene_id, prmt5_gripps$gene_id)),
  mart = mart
)

biotypes$Group <- ifelse(biotypes$ensembl_gene_id %in% gripps$gene_id, "GRIPPs", "PRMT5-GRIPPs")

# Collapse gene types for display
summary_df <- biotypes %>%
  dplyr::mutate(
    RNA_Class = case_when(
      grepl("pseudogene", gene_biotype, ignore.case = TRUE) ~ "Pseudogene",
      gene_biotype == "TEC" ~ "TEC",
      gene_biotype == "lncRNA" ~ "lncRNA",
      gene_biotype == "protein_coding" ~ "protein_coding",
      TRUE ~ "other"
    )
  ) %>%
  dplyr::group_by(Group, RNA_Class) %>%
  dplyr::summarise(Count = dplyr::n()) %>%
  dplyr::mutate(Percentage = 100 * Count / sum(Count))

# Generate the stacked bar plot
S

```