---
title: "GRIPP Definition"
output:
  pdf_document: default
  html_document:
    df_print: paged
author: "David Shechter, Maxim Maron, and Joe DeAngelo"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

###Introduction

*This analysis refines the definition of GRIPPs by integrating: 1. Chromatin-retained genes (Fig 6) 2. Retained introns (Fig 7) 3. Accounting for multiple retained introns per gene using summary metrics*

This analysis requires STAR mapping to a joint hg38-dm6 genome and a pre-built gtf with "dm6" in the Chromsome column for any control spike-in gene. These files are available at [www.github.com/shechterlab](www.github.com/shechterlab)

```{r setup, include=FALSE}
library(tidyverse)
library(data.table)

# Set the WD here
manuscript_dir <- 'C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/'
working_dir <- 'C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 7/GRIPPs/'
knitr::opts_knit$set(root.dir = normalizePath(working_dir))
setwd(working_dir)
getwd()

# Create output directory
output_dir <- paste0(working_dir, "Output/")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir)
  }

```


```{r load and process data}

# Load chromatin-enriched genes (Fig 6)
chromatin_data <- read.csv(paste0(manuscript_dir, "Figure 6/Output/GSK591_chromatin_vs_DMSO_chromatin_SpikeInNormalized_DESEQ2.csv"))

chromatin_genes <- chromatin_data %>%
  filter(padj < 0.05 & log2FoldChange > 0.58) %>%
  dplyr::select(GeneSymbol, log2FoldChange, padj) %>%
  mutate(Source = "Chromatin_Enriched")

# Load retained introns (Fig 7)
intron_data <- read.csv(paste0(manuscript_dir, "Figure 7/Output/GSK591_Chromatin_vs_DMSO_Chromatin_DESeq2_results.csv"))

intron_data <- intron_data %>%
  mutate(GeneSymbol = sub(":.*", "", IntronID))  # Extract GeneSymbol from IntronID

# Load intron and exon counts
intron_counts <- fread(paste0(manuscript_dir, "Figure 7/featureCounts_IntronsExons/Intron_Count_hg38_dm6_A549_frac_PRMTi_s2.txt")) %>%
  mutate(GeneSymbol = sub(":.*", "", Geneid)) %>%
  group_by(GeneSymbol) %>%
  summarise(Total_Intron_Reads = sum(rowSums(across(starts_with("DMSO") | starts_with("GSK591")))))

exon_counts <- fread(paste0(manuscript_dir, "Figure 7/featureCounts_IntronsExons/Exon_Count_hg38_dm6_A549_frac_PRMTi_s2.txt")) %>%
  mutate(GeneSymbol = sub(":.*", "", Geneid)) %>%
  group_by(GeneSymbol) %>%
  summarise(Total_Exon_Reads = sum(rowSums(across(starts_with("DMSO") | starts_with("GSK591")))))

# Merge and calculate IntronExon_Ratio
gene_counts <- intron_counts %>%
  inner_join(exon_counts, by = "GeneSymbol") %>%
  mutate(IntronExon_Ratio = Total_Intron_Reads / Total_Exon_Reads)

# Recalculate IntronExon_Ratio (if missing)
count_ratios <- intron_counts %>%
  inner_join(exon_counts, by = "GeneSymbol") %>%
  mutate(IntronExon_Ratio = Total_Intron_Reads / Total_Exon_Reads)

# Merge counts with intron data
intron_data <- intron_data %>%
  inner_join(count_ratios, by = "GeneSymbol")

# Filter significantly retained introns
retained_introns <- intron_data %>%
  filter(padj < 0.1 & log2FoldChange > 0.5)

# Summarize introns at the gene level
intron_summary <- retained_introns %>%
  group_by(GeneSymbol) %>%
  summarise(
    Avg_Intron_FC = mean(log2FoldChange),
    Median_Intron_FC = median(log2FoldChange),
    Avg_IntronExon_Ratio = mean(IntronExon_Ratio),
    Median_IntronExon_Ratio = median(IntronExon_Ratio),
    Num_Retained_Introns = n_distinct(IntronID)
  )

```

#GRIPPS definition

##Chromatin Retention as a Key Criterion

*Chromatin-associated genes are an essential component of the GRIPP definition. Using log2FoldChange (scaled or unscaled) ensures that only genes with strong chromatin enrichment contribute significantly to the GRIPP score.*

##Normalization Across Genes:

*z-score scaled Chromatin FC ensures that the metric is standardized, allowing fair comparison across genes, even if their absolute fold-change values differ greatly. Without scaling, highly expressed genes with moderate fold changes might dominate.*

##Combined Metric:

*By combining Chromatin FC with Avg_IntronExon_Ratio, you ensure that genes meeting both criteria (strong chromatin enrichment and significant intron retention) are emphasized. This helps filter out false positives where only one feature is significant.*


```{r define GRIPPs}


# Merge with chromatin-enriched genes
combined_data <- chromatin_genes %>%
  inner_join(intron_summary, by = "GeneSymbol")


# Scale metrics (by Z-score) and calculate GRIPP_Score
combined_data <- combined_data %>%
  mutate(
    Scaled_Chromatin_FC = (log2FoldChange - mean(log2FoldChange)) / sd(log2FoldChange),
    Scaled_Intron_FC = (Avg_Intron_FC - mean(Avg_Intron_FC)) / sd(Avg_Intron_FC),
    Scaled_IntronExon_Ratio = (Avg_IntronExon_Ratio - mean(Avg_IntronExon_Ratio)) / sd(Avg_IntronExon_Ratio),
    GRIPP_Score = sqrt(Scaled_Chromatin_FC * (5 + Scaled_IntronExon_Ratio)), #geometric mean
    GRIPP_Definition = GRIPP_Score > 0.1  # Threshold
  )

#sanity check, print histograms

hist(combined_data$Scaled_Chromatin_FC, main = "Distribution of zscore Scaled Chromatin FC", xlab = "Scaled Chromatin FC", plot = TRUE)

hist(combined_data$Scaled_IntronExon_Ratio, main = "Distribution of Scaled Intron FC", xlab = "Scaled Intron FC", plot = TRUE)


#Method 2 - rank analysis
# Rank metrics
#combined_data <- combined_data %>%
#  mutate(
#    Rank_Chromatin_FC = rank(-log2FoldChange),  # Higher FC gets higher rank
#    Rank_IntronExon_Ratio = rank(-Avg_IntronExon_Ratio),  # Higher ratio gets higher rank
#    Rank_Intron_FC = rank(-Avg_Intron_FC)  # Higher FC gets higher rank (optional)
#  )

# Combine ranks (adjust weights as needed)
#combined_data <- combined_data %>%
#  mutate(
#    Composite_Rank = 0.5 * Rank_Chromatin_FC + 0.5 * Rank_Intron_FC,  # Weighted sum
#    GRIPP_Definition = Composite_Rank <= quantile(Composite_Rank, 0.25)  # Top 10% as GRIPPs
#  )

# Filter for GRIPPs
#gripps <- combined_data %>%
 # filter(GRIPP_Definition) %>%
#  arrange(Composite_Rank)




# Filter for GRIPPs
gripps <- combined_data %>%
  filter(GRIPP_Definition) %>%
  arrange(desc(GRIPP_Score))

# Save output
write.csv(gripps, file = paste0(output_dir, "GRIPPs_Updated.csv"), row.names = FALSE)

# Summarize GRIPPs
summary_table <- gripps %>%
  summarise(
    Total_GRIPPs = n(),
    Mean_GRIPP_Score = mean(GRIPP_Score),
    Median_GRIPP_Score = median(GRIPP_Score)
  )
print(summary_table)


```

```{r visualize GRIPPs}


# Bar plot of GRIPP metrics
ggplot(gripps, aes(x = reorder(GeneSymbol, -GRIPP_Score), y = GRIPP_Score)) +
  geom_bar(stat = "identity", fill = "#2D8942") +
  theme_minimal() +
  labs(
    title = "GRIPP Scores by Gene",
    x = "Gene ID",
    y = "GRIPP Score"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
```{r scatter plots}

# Scatter plot for Chromatin FC Rank vs Intron FC Rank
plot_gene_vs_introns <- ggplot(combined_data, aes(x = Rank_Chromatin_FC, y = Rank_Intron_FC)) +
  geom_point(alpha = 0.6, color = "#2D86B2") +
  theme_minimal() +
  labs(
    title = "Scatter Plot: Chromatin FC Rank vs Intron FC Rank",
    x = "Chromatin FC Rank",
    y = "Intron FC Rank"
  ) +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

# Print the plot
print(plot_gene_vs_introns)


# Scatter plot for Chromatin FC Rank vs Intron FC Rank
plot_gene_vs_intronExonRatio <- ggplot(combined_data, aes(x = Rank_Chromatin_FC, y = Rank_IntronExon_Ratio)) +
  geom_point(alpha = 0.6, color = "#2D8942") +
  theme_minimal() +
  labs(
    title = "Scatter Plot: Chromatin FC Rank vs IntronExonRatio Rank",
    x = "Chromatin FC Rank",
    y = "IntronExon Rank"
  ) +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

# Print the plot
print(plot_gene_vs_intronExonRatio)


 # Scatter plot for Chromatin log2FC vs Avg Intron log2FC
> plot_log2fc <- ggplot(combined_data, aes(x = log2FoldChange, y = Avg_Intron_FC)) +
+     geom_point(alpha = 0.6, color = "#2D86B2") +
+     theme_minimal() +
+     labs(
+         title = "Scatter Plot: Chromatin log2FC vs Intron log2FC",
+         x = "Chromatin log2 Fold Change (log2FC)",
+         y = "Average Intron log2 Fold Change (Avg_Intron_FC)"
+     ) +
+     geom_smooth(method = "lm", color = "red", se = FALSE) +  # Optional: Add a trendline
+     theme(plot.title = element_text(hjust = 0.5))  # Center the title
> 
> # Print the plot
> print(plot_log2fc)


```


```{r intronexon ratios for fractions}
# Load intron counts, skipping the first comment line
intron_counts <- fread(
  paste0(manuscript_dir, "Figure 7/featureCounts_IntronsExons/Intron_Count_hg38_dm6_A549_frac_PRMTi_s2.txt"),
  skip = 1
) %>%
  mutate(GeneSymbol = sub(":.*", "", Geneid)) %>%
  group_by(GeneSymbol) %>%
  summarise(
    Chromatin_Intron_Reads = rowSums(across(starts_with("Chromatin"))),
    Nucleoplasm_Intron_Reads = rowSums(across(starts_with("Nucleoplasm"))),
    Cytoplasm_Intron_Reads = rowSums(across(starts_with("Cytoplasm")))
  )

# Load exon counts, skipping the first comment line
exon_counts <- fread(
  paste0(manuscript_dir, "Figure 7/featureCounts_IntronsExons/Exon_Count_hg38_dm6_A549_frac_PRMTi_s2.txt"),
  skip = 1
) %>%
  mutate(GeneSymbol = sub(":.*", "", Geneid)) %>%
  group_by(GeneSymbol) %>%
  summarise(
    Chromatin_Exon_Reads = rowSums(across(starts_with("Chromatin"))),
    Nucleoplasm_Exon_Reads = rowSums(across(starts_with("Nucleoplasm"))),
    Cytoplasm_Exon_Reads = rowSums(across(starts_with("Cytoplasm")))
  )

# Merge intron and exon counts for each fraction
fraction_ratios <- intron_counts %>%
  inner_join(exon_counts, by = "GeneSymbol") %>%
  mutate(
    Chromatin_IntronExon_Ratio = Chromatin_Intron_Reads / Chromatin_Exon_Reads,
    Nucleoplasm_IntronExon_Ratio = Nucleoplasm_Intron_Reads / Nucleoplasm_Exon_Reads,
    Cytoplasm_IntronExon_Ratio = Cytoplasm_Intron_Reads / Cytoplasm_Exon_Reads
  )

# Filter genes with valid ratios (to avoid division by zero)
fraction_ratios <- fraction_ratios %>%
  filter(is.finite(Chromatin_IntronExon_Ratio),
         is.finite(Nucleoplasm_IntronExon_Ratio),
         is.finite(Cytoplasm_IntronExon_Ratio))

# Visualize chromatin vs nucleoplasm ratios
ggplot(fraction_ratios, aes(x = Chromatin_IntronExon_Ratio, y = Nucleoplasm_IntronExon_Ratio)) +
  geom_point(alpha = 0.5, color = "blue") +
  theme_minimal() +
  labs(
    title = "Chromatin vs Nucleoplasm: Per-Gene Intron/Exon Ratios",
    x = "Chromatin Intron/Exon Ratio",
    y = "Nucleoplasm Intron/Exon Ratio"
  ) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red")

# Visualize chromatin vs cytoplasm ratios
ggplot(fraction_ratios, aes(x = Chromatin_IntronExon_Ratio, y = Cytoplasm_IntronExon_Ratio)) +
  geom_point(alpha = 0.5, color = "green") +
  theme_minimal() +
  labs(
    title = "Chromatin vs Cytoplasm: Per-Gene Intron/Exon Ratios",
    x = "Chromatin Intron/Exon Ratio",
    y = "Cytoplasm Intron/Exon Ratio"
  ) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red")

```