---
title: "A549 Intron/Exon analysis, DESeq2 Analysis with hg38 and dm6 spike-in"
output:
  pdf_document: default
  html_document:
    df_print: paged
author: "David Shechter, Maxim Maron, and Joe DeAngelo"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

*This R Markdown script is a comprehensive analysis pipeline for comparing intron and exon RNA-seq data across A549 cell fractionation and the effects of PRMT5 inhibition (GSK591). The pipeline integrates hg38 (human genome) and dm6 (Drosophila spike-in controls) for normalization*

This analysis requires STAR mapping to a joint hg38-dm6 genome and a pre-built gtf with "dm6" in the Chromsome column for any control spike-in gene. These files are available at [www.github.com/shechterlab](www.github.com/shechterlab)

```{r setup, include=FALSE}

# Load necessary libraries
library(tidyverse)
library(data.table)
library(DESeq2)
library(ggrepel)
library(biomaRt) # for intron position analysis
library(dplyr)
library(ggpointdensity)
library(viridis)

# Set the WD here
working_dir <- 'C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 7/'
knitr::opts_knit$set(root.dir = normalizePath(working_dir))
setwd(working_dir)
getwd()

# Create output directory
output_dir <- paste0(working_dir, "Output/")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir)
  }

# Define a function to read and process the data
process_data <- function(file_path, type) {
  data <- fread(file_path)
  
  # Extract the gene name and IDs
  data <- data %>%
    mutate(GeneName = sub(":.*", "", Geneid),
           ExonID = case_when(type == "exon" ~ Geneid, TRUE ~ NA_character_),
           IntronID = case_when(type == "intron" ~ Geneid, TRUE ~ NA_character_)) %>%
    dplyr::select(-Geneid)
  # Apply cutoff of 20 counts in any 3 samples (e.g. should be truly present in at least one fraction and condition)
  data <- data %>%
    dplyr::filter(rowSums(dplyr::select(., starts_with("DMSO") | starts_with("GSK591")) >= 20) >= 3)
  return(data)
}

# Read the data
intron_data <- process_data(paste0(working_dir,'featureCounts_IntronsExons/Intron_Count_hg38_dm6_A549_frac_PRMTi_s2.txt'), "intron")
exon_data <- process_data(paste0(working_dir,'featureCounts_IntronsExons/Exon_Count_hg38_dm6_A549_frac_PRMTi_s2.txt'), "exon")


# Read hg38 intron lengths, although this won't actually be used as it is not limited to expressed genes
#hg38_intron_lengths <- read.csv("C:/Users/david/OneDrive/Bioinformatics/PRMTi/2024-05_RNAseq_PRMT5i-Fractionation/IntronExonAnalysis/hg38dm6-SpikeInAnalysis/hg38_gencode_intron_lengths.txt", header = FALSE, col.names = c("Length"))


```

```{r Create gene summary table}

# Create a summary table for genes
create_gene_summary <- function(exon_data, intron_data) {
  # Actual reads per condition/replicate
  exon_long <- exon_data %>%
    pivot_longer(cols = starts_with("DMSO") | starts_with("GSK591"), names_to = "Condition_Replicate", values_to = "ExonReadCount")
  
  intron_long <- intron_data %>%
    pivot_longer(cols = starts_with("DMSO") | starts_with("GSK591"), names_to = "Condition_Replicate", values_to = "IntronReadCount")
  
  # Count the number of exons and introns per gene
  exon_count <- exon_data %>%
    group_by(GeneName) %>%
    summarize(ExonCount = n(), .groups = 'drop')
  
  intron_count <- intron_data %>%
    group_by(GeneName) %>%
    summarize(IntronCount = n(), .groups = 'drop')
  
  # Average reads per condition
  exon_avg <- exon_long %>%
    separate(Condition_Replicate, into = c("Condition", "Replicate"), sep = "_rep") %>%
    group_by(GeneName, Condition) %>%
    summarize(avgExon_reads = mean(ExonReadCount), .groups = 'drop')
  
  intron_avg <- intron_long %>%
    separate(Condition_Replicate, into = c("Condition", "Replicate"), sep = "_rep") %>%
    group_by(GeneName, Condition) %>%
    summarize(avgIntron_reads = mean(IntronReadCount), .groups = 'drop')
  
  # Join exon and intron summaries
  gene_summary <- inner_join(exon_avg, intron_avg, by = c("GeneName", "Condition"))
  
  # Add exon and intron counts
  gene_summary <- gene_summary %>%
    left_join(exon_count, by = "GeneName") %>%
    left_join(intron_count, by = "GeneName")
  
  # Calculate IntronExon_Ratio for each condition, add 1 to make sure there is no divide by 0
  gene_summary <- gene_summary %>%
    mutate(IntronExon_Ratio = avgIntron_reads / (avgExon_reads + 1))
  
  return(gene_summary)
}

gene_summary <- create_gene_summary(exon_data, intron_data)

# Split condition into Treatment and Fraction
gene_summary <- gene_summary %>%
  separate(Condition, into = c("Treatment", "Fraction"), sep = "_")

# Ensure the order of levels for the Fraction factor
gene_summary$Fraction <- factor(gene_summary$Fraction, levels = c("Chromatin", "Nucleoplasm", "Cytoplasm"))

 # Save the gene summary table to a CSV file
write.csv(gene_summary, file = paste0(output_dir, 'A549-Fractionation_IntronExon-featureCounts_gene_summary.csv'), row.names = FALSE)


```

```{r Create intron summary table}

# Create a summary table for introns
create_intron_summary <- function(intron_data) {
  # Pivot the data to long format
  intron_long <- intron_data %>%
    pivot_longer(cols = starts_with("DMSO") | starts_with("GSK591"), names_to = "Condition_Replicate", values_to = "IntronReadCount") %>%
    separate(Condition_Replicate, into = c("Condition", "Replicate"), sep = "_rep")

  # Calculate average reads per condition for each intron
  intron_summary <- intron_long %>%
    group_by(IntronID, Condition) %>%
    summarize(avgIntron_reads = mean(IntronReadCount), Length = dplyr::first(Length), .groups = 'drop')


  return(intron_summary)
}

intron_summary <- create_intron_summary(intron_data)

write.csv(intron_summary, file = paste0(output_dir, 'A549-Fractionation_IntronExon-featureCounts_intron_summary.csv'), row.names = FALSE)

# Create an "all" table from all featureCount expressed genes (cutoff > 20 counts)
all_intron_lengths <- intron_data %>%
  dplyr::select(IntronID, Length) %>%
  dplyr::distinct() %>%
  dplyr::mutate(Fraction = "All_featureCounts")


```

***Violin plot of intron/exon ratio, Per Condition ***

```{r violin plot of intron/exon ratio per condition}

# Ensure the order of levels for the Fraction factor in the gene_summary data frame
gene_summary$Fraction <- factor(gene_summary$Fraction, levels = c("Chromatin", "Nucleoplasm", "Cytoplasm"))

# Apply natural log transformation to the intron/exon ratio
gene_summary <- gene_summary %>%
  dplyr::filter(!grepl("dm6", GeneName)) %>%
  dplyr::mutate(Log_IntronExon_Ratio = log1p(IntronExon_Ratio))

# Perform KS test for each fraction between DMSO and GSK591
ks_results <- gene_summary %>%
  group_by(Fraction) %>%
  summarize(ks_p_value = ks.test(Log_IntronExon_Ratio[Treatment == "DMSO"],
                                 Log_IntronExon_Ratio[Treatment == "GSK591"])$p.value)

fraction_results <- gene_summary %>%
  group_by(Fraction) %>%
  summarize(
    ks_p_value = ks.test(Log_IntronExon_Ratio[Treatment == "DMSO"],
                         Log_IntronExon_Ratio[Treatment == "GSK591"])$p.value,
    ks_D = ks.test(Log_IntronExon_Ratio[Treatment == "DMSO"],
                   Log_IntronExon_Ratio[Treatment == "GSK591"])$statistic,
    wilcox_p_value = wilcox.test(Log_IntronExon_Ratio[Treatment == "DMSO"],
                                 Log_IntronExon_Ratio[Treatment == "GSK591"])$p.value,
    wilcox_statistic = wilcox.test(Log_IntronExon_Ratio[Treatment == "DMSO"],
                                   Log_IntronExon_Ratio[Treatment == "GSK591"])$statistic
  )

# Merge KS test results with the gene_summary for labeling
gene_summary <- gene_summary %>%
  left_join(ks_results, by = "Fraction")

# Reverse the levels of Fraction to place Chromatin at the bottom
gene_summary$Fraction <- factor(gene_summary$Fraction, levels = c("Cytoplasm", "Nucleoplasm", "Chromatin"))

# Adjust the order of Treatment within each group
gene_summary <- gene_summary %>%
  mutate(Treatment = factor(Treatment, levels = c("DMSO", "GSK591"))) %>%
  arrange(Fraction, Treatment)

# Create the horizontal violin plot
violin_plot <- ggplot(gene_summary, aes(y = interaction(Fraction, Treatment, lex.order = TRUE), 
                                        x = Log_IntronExon_Ratio, fill = Treatment)) +
  geom_violin(trim = FALSE, scale = "width", orientation = "y") +
  geom_boxplot(width = 0.1, position = position_dodge(0.9), outlier.shape = NA) +
  scale_fill_manual(values = c("DMSO" = "#AAAAAA", "GSK591" = "#2D8942")) +
  theme_minimal() +
  labs(title = "Natural Log-Transformed Distribution of Per Gene Intron/Exon Ratio per Condition and Fraction",
       y = "Fraction and Treatment",
       x = "ln(Intron/Exon Ratio + 1)") +
  geom_text(data = fraction_results, aes(y = 1.5, x = 3,
                                         label = paste0("Wilcoxon p-value: ", format(wilcox_p_value, digits = 2))),
            inherit.aes = FALSE, size = 2, color = "black") +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),  # Centered Title
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  ) +
  xlim(-0.1, 3)



print(violin_plot)

# Save as PNG
ggsave(paste0(output_dir, 'ln-IntronExon-Ratio_Violin_Plot.png'), plot = violin_plot, width = 10, height = 7)

# Save as PDF
ggsave(paste0(output_dir, 'ln-IntronExon-Ratio_Violin_Plot.pdf'), plot = violin_plot, width = 10, height = 7)


```


```{r Volcano Plot Function, INCLUDE = FALSE}

# Function to create volcano plots
create_volcano_plot <- function(data, title) {
  data$log10padj <- -log10(data$padj)
  
  max_abs_x <- max(abs(data$log2FoldChange)) * 1.4 # give 40% buffer on axes
  max_y <- max(data$log10padj) * 1.4 # 40% buffer
  
  p <- ggplot(data, aes(x = log2FoldChange, y = log10padj)) +
    geom_point(color = "#BEBEBE", size = 1.5, alpha = 1, shape = 16) +
    geom_text_repel(data = subset(data, log10padj > 3 & log2FoldChange > 0.58),
                    aes(label = GeneSymbol),
                    size = 3,
                    colour = "#B2182B",
                    segment.color = "#B2182B",
                    max.overlaps = 7,
                    box.padding = 0.5,
                    force = 20,
                    min.segment.length = 0,
                    nudge_x = 0.1,
                    direction = "both") +
    geom_text_repel(data = subset(data, log10padj > 3 & log2FoldChange < -0.58),
                    aes(label = GeneSymbol),
                    size = 3,
                    colour = "#2166AC",
                    segment.color = "#2166AC",
                    max.overlaps = 7,
                    box.padding = 0.5,
                    force = 10,
                    min.segment.length = 0,
                    nudge_x = -0.2,
                    direction = "both") +
    geom_point(data = . %>% filter(log10padj > 1.301 & log2FoldChange > 0.58),
               size = 2, color = "#B2182B") +
    geom_point(data = . %>% filter(log10padj > 1.301 & log2FoldChange < -0.58),
               size = 2, color = "#2166AC") +
    geom_vline(xintercept = 0.58, color = "#B2182B", linetype = "dotdash", linewidth = 0.5) +
    geom_vline(xintercept = -0.58, color = "#2166AC", linetype = "dotdash", linewidth = 0.5) +
    geom_vline(xintercept = 0, color = "black", linetype = "solid", linewidth = 0.5) +
    geom_hline(yintercept = 1.301, color = "grey", linetype = "dotdash", linewidth = 0.5) +
    theme(aspect.ratio = 1) +
    scale_x_continuous(limits = c(-max_abs_x, max_abs_x)) +
    scale_y_continuous(limits = c(0, max_y)) +
    theme(
      legend.position = "none",
      axis.text.y = element_text(size = 12, colour = "black"),
      axis.text.x = element_text(size = 12, colour = "black"),
      axis.title.y = element_text(size = 12, colour = "black"),
      axis.title.x = element_text(size = 12, colour = "black"),
      panel.background = element_blank(),
      panel.grid.major = element_line(color = "#D3D3D380", linewidth = 0.5),
      panel.grid.minor = element_line(color = "#D3D3D380", linewidth = 0.25),
      axis.line = element_line(colour = "black"),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
      plot.title = element_text(size = 14, hjust = 0)
    ) +
    labs(title = title,
         x = expression(atop("log"[2] * " Fold Change", "Comparison")),
         y = expression('-log'[10]*' padj'))
  
  return(p)
}

```

```{r DEseq2 setup and spike-in normalization}

#Prepare data for DESeq2 analysis. Select only the introns that are not dm6 (e.g. hg38)
hg38_counts <- intron_data %>%
  dplyr::filter(!grepl("dm6", IntronID)) %>%
  dplyr::select(-Chr, -Start, -End, -Strand, -Length, -GeneName, -ExonID) %>%
  column_to_rownames(var = "IntronID")

# Extract dm6 spike-in counts for normalization
dm6_counts <- intron_data %>%
  dplyr::filter(grepl("dm6", IntronID)) %>%
  dplyr::select(-Chr, -Start, -End, -Strand, -Length, -GeneName, -ExonID) %>%
  column_to_rownames(var = "IntronID")

#  Merge hg38 and dm6 counts
all_counts <- rbind(hg38_counts, dm6_counts)

# Check the number of columns and define conditions accordingly
num_samples <- ncol(all_counts)
conditions <- rep(c("DMSO_Chromatin", "DMSO_Nucleoplasm", "DMSO_Cytoplasm",
                    "GSK591_Chromatin", "GSK591_Nucleoplasm", "GSK591_Cytoplasm"), each = 3)
if (length(conditions) != num_samples) {
  stop("The number of conditions does not match the number of samples.")
}

# Prepare colData for DESeq2
colData <- data.frame(row.names = colnames(all_counts), condition = factor(conditions))

# Create DESeq2 dataset including spike-in controls
dds <- DESeqDataSetFromMatrix(countData = all_counts, colData = colData, design = ~ condition)

# Calculate size factors using spike-in controls
spike_in <- grepl("dm6", rownames(dds))
dds <- estimateSizeFactors(dds, controlGenes = spike_in)

# Print size factors
print(sizeFactors(dds))

# Remove spike-in controls from the dataset
dds <- dds[!spike_in, ]

# Extract feature lengths
feature_lengths <- intron_data %>%
  dplyr::select(IntronID, Length) %>%
  dplyr::distinct()
lengths <- feature_lengths$Length
names(lengths) <- feature_lengths$IntronID

# Ensure lengths vector matches rows of dds
lengths <- lengths[match(rownames(dds), names(lengths))]

# Create offset matrix
offset_matrix <- matrix(log(rep(lengths, each = ncol(dds))), nrow = nrow(dds), byrow = FALSE)
rownames(offset_matrix) <- rownames(dds)
colnames(offset_matrix) <- colnames(dds)

# Add the offset matrix to the DESeq2 dataset
assay(dds, "offsets") <- offset_matrix

# Perform DESeq2 analysis using the offsets
dds <- DESeq(dds, test = "LRT", reduced = ~ 1)

# Extract normalized counts, retain IntronID as a column
norm_counts <- as.data.frame(counts(dds, normalized = TRUE)) %>%
  tibble::rownames_to_column(var = "IntronID")

# Save normalized counts
write.csv(norm_counts, paste0(output_dir, 'A549-GSK591andFractionation_dm6-Normalized_hg38-INTRON-counts_DESeq2.csv'), row.names = FALSE)

```

```{r DESeq2 function call for intron analysis with spike in, also intron feature analysis}

# Function for DESeq2 Analysis for Intron Data, use pre-exisitng sizeFactors and Dispersion estimates from above chunk

# Function for DESeq2 Analysis for Intron Data
run_deseq2_intron_analysis <- function(dds, test_condition, reference_condition, output_prefix) {
  
  #add working dir to outputprefix
  output_prefix <- paste0(output_dir, output_prefix) 
  
  # Subset DESeq2 dataset for the specified conditions
  dds_subset <- dds[, dds$condition %in% c(test_condition, reference_condition)]
  dds_subset$condition <- droplevels(dds_subset$condition)
  
  # Use pre-calculated size factors
  sizeFactors(dds_subset) <- sizeFactors(dds)[colnames(dds_subset)]
  
  # Print the size factors for confirmation
  print("Size factors being used:")
  print(sizeFactors(dds_subset))
  
  # Perform DESeq2 analysis including dispersion estimation
  dds_subset <- DESeq(dds_subset)
  
  # Extract results for the specific comparison
  res <- results(dds_subset, contrast = c("condition", test_condition, reference_condition))
  res <- res[!is.na(res$padj), ]
  
  # Save the results to a CSV file, converting rowname to column
  write.csv(
    (as.data.frame(res) %>%
    tibble::rownames_to_column(var = "IntronID")
    ), paste0(output_prefix, "_DESeq2_results.csv"), row.names = FALSE)
  
  # Create a volcano plot
  volcano_data <- as.data.frame(res) %>%
    tibble::rownames_to_column("IntronID") %>%
    mutate(log10padj = -log10(padj), GeneSymbol = IntronID)
  
  volcano_plot <- create_volcano_plot(volcano_data, paste0(test_condition, " vs ", reference_condition))
  print(volcano_plot)
  

  # Save the volcano plot
  ggsave(paste0(output_prefix, "_Volcano_Plot.png"), plot = volcano_plot, width = 10, height = 7)
  ggsave(paste0(output_prefix, "_Volcano_Plot.pdf"), plot = volcano_plot, width = 10, height = 7)
  
  return(res)
}

######################################################################

# Function to create volcano plots
create_volcano_plot <- function(data, title) {
  ggplot(data, aes(x = log2FoldChange, y = log10padj)) +
    geom_point(aes(color = log10padj > 2 & abs(log2FoldChange) > 1), size = 1.5) +
    scale_color_manual(values = c("grey", "red")) +
    theme_minimal() +
    labs(title = title, x = "Log2 Fold Change", y = "-Log10 Adjusted p-value") +
    theme(legend.position = "none")
}

#####################################################################

# Function to analyze intron enrichment and plot histogram
analyze_intron_enrichment <- function(res_data, intron_data, output_name, padj_cutoff = 0.05, log2fc_cutoff = 0.58) {
  
  #add wd to outputname
  output_name <- paste0(output_dir, output_name) 
  
  # Convert DESeqResults to a data frame for dplyr compatibility, ensure that IntronID is propagated to res_df
    res_df <- as.data.frame(res_data) %>%
  tibble::rownames_to_column(var = "IntronID")
  
  # Filter for significant introns
  sig_introns <- res_df %>%
    dplyr::filter(padj < padj_cutoff & abs(log2FoldChange) > log2fc_cutoff)
  
  # Extract gene names and count the number of significant introns per gene
  sig_introns <- sig_introns %>%
    dplyr::mutate(GeneName = sub(":.*", "", IntronID))
  
  sig_intron_counts <- sig_introns %>%
    group_by(GeneName) %>%
    summarize(SigIntronCount = n())
  
  # Extract total intron counts per gene from intron_data
  total_intron_counts <- intron_data %>%
    dplyr::filter(!grepl("dm6", IntronID)) %>%
    group_by(GeneName) %>%
    summarize(TotalIntronCount = n())
  
  # Merge the significant intron counts with total intron counts
  enrichment_data <- left_join(sig_intron_counts, total_intron_counts, by = "GeneName")
  
  # Calculate the percentage of enriched introns per gene
  enrichment_data <- enrichment_data %>%
    dplyr::mutate(PercentageEnrichedIntrons = (SigIntronCount / TotalIntronCount) * 100)
  
  # Replace NA values with 0 (genes with no significant introns)
  enrichment_data$PercentageEnrichedIntrons[is.na(enrichment_data$PercentageEnrichedIntrons)] <- 0
  
  # Plot histogram of percentage of enriched introns per gene
  intron_percentage_histogram <- ggplot(enrichment_data, aes(x = PercentageEnrichedIntrons)) +
    geom_histogram(binwidth = 2, fill = "gray", color = "darkgray") +
    theme_minimal() +
    xlim(0, 100) +
    labs(title = paste(output_name, "Percentage of Enriched Introns per Gene"),
         x = "Percentage of Enriched Introns",
         y = "Number of Genes")
  
  print(intron_percentage_histogram)

  
  # Save the histogram
  ggsave(paste0(output_name, "_Percentage_Enriched_Introns_Histogram.png"), plot = intron_percentage_histogram, width = 10, height = 3)
  ggsave(paste0(output_name, "_Percentage_Enriched_Introns_Histogram.pdf"), plot = intron_percentage_histogram, width = 10, height = 3)
  
  return(enrichment_data)
}

########################################################################

# Function to extract significantly enriched introns and perform length analysis

analyze_intron_length <- function(res_data, intron_data, output_prefix, fraction_name) {
  # Convert DESeqResults to a data frame for dplyr compatibility, ensure that IntronID is propagated to res_df
  res_df <- as.data.frame(res_data)%>%
  tibble::rownames_to_column(var = "IntronID")
  
  # Function to extract significantly enriched introns for a given fraction, remove dm6 introns
  extract_sig_introns <- function(res_data, intron_data, fraction_name) {
    res_data %>%
     dplyr::filter(padj < 0.05 & log2FoldChange > 0.58) %>%
      dplyr::select(IntronID) %>%
      inner_join(intron_data, by = "IntronID") %>%
      filter(!grepl("dm6", IntronID)) %>%
      mutate(Fraction = fraction_name)
  }

  # Extract significantly enriched introns for the specified fraction
  sig_introns <- extract_sig_introns(res_df, intron_data, fraction_name)

# Filter for expressed genes based on exon data, at least 10 counts in 1 sample
expressed_genes <- exon_data %>%
  dplyr::filter(rowSums(dplyr::select(., starts_with("DMSO") | starts_with("GSK591")) > 10) > 1) %>%
  dplyr::pull(GeneName) %>%
  unique()

# Filter intron data to include only introns from expressed genes
all_intron_lengths <- intron_data %>%
  dplyr::filter(GeneName %in% expressed_genes) %>%
  dplyr::select(IntronID, Length) %>%
  filter(!grepl("dm6", IntronID)) %>%
  distinct() %>%
  mutate(Fraction = "All_featureCounts")

  # Filter out non-finite lengths
  sig_introns <- sig_introns %>% filter(is.finite(Length))
  all_intron_lengths <- all_intron_lengths %>% filter(is.finite(Length))

  # Combine data for "All_featureCounts" and the specified fraction
  plot_data <- bind_rows(
    sig_introns %>% dplyr::select(Length, Fraction),
    all_intron_lengths %>% dplyr::select(Length, Fraction)
  )

  # Set the order of the fractions
  plot_data$Fraction <- factor(plot_data$Fraction, levels = c("All_featureCounts", fraction_name))

  # Calculate the medians for each fraction
  medians <- plot_data %>%
    group_by(Fraction) %>%
    summarise(median_length = median(Length))

  # Calculate median differences compared to "All_featureCounts"
  median_differences <- medians %>%
    dplyr::filter(Fraction != "All_featureCounts") %>%
    dplyr::mutate(median_diff = median_length - medians$median_length[medians$Fraction == "All_featureCounts"])

  # Perform KS and Wilcoxon tests for the fraction against all featureCounts introns
  test_results <- plot_data %>%
    dplyr::filter(Fraction == fraction_name) %>%
    summarise(
      ks_p_value = ks.test(log10(Length), log10(all_intron_lengths$Length))$p.value,
      wilcox_p_value = wilcox.test(log10(Length), log10(all_intron_lengths$Length))$p.value
    )

# Calculate the count of introns for each fraction
intron_counts <- plot_data %>%
  group_by(Fraction) %>%
  summarise(n_introns = n(), .groups = "drop")

# Extract counts for specific fractions
n_all <- intron_counts %>% filter(Fraction == "All_featureCounts") %>% pull(n_introns)
n_test <- intron_counts %>% filter(Fraction == fraction_name) %>% pull(n_introns)


  # Create a violin plot of the log10 lengths of significantly enriched introns with Wilcoxon p-values and median differences
violin_length_plot <- ggplot(plot_data, aes(y = Fraction, x = log10(Length), fill = Fraction)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7, color = NA) +  # Horizontal violin plot
  geom_boxplot(width = 0.1, position = position_dodge(0.9), outlier.shape = NA, color = "black", orientation = "y") +  # Box plot overlay
  stat_summary(fun = median, geom = "point", shape = 18, size = 3, color = "black") +  # Add median points
  # Add annotations for count and median
  stat_summary(
    fun.data = function(x) {
      data.frame(
        x = max(log10(plot_data$Length), na.rm = TRUE) * 1.05,
        label = paste0("count = ", length(x), "\nmedian = ", round(median(log10(x)), 3))
      )
    },
    geom = "text",
    position = position_dodge(width = 0.9),
    size = 3,
    aes(label = ..label..)
  ) +
  scale_fill_manual(values = c("All_ExpressedGeneIntrons" = "gray", fraction_name = "#2D8942")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, color = "black"),
    axis.title.x = element_text(size = 10, color = "black"),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 10, color = "black"),
    legend.position = "none",
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  labs(
    title = paste("Length of All Expressed Gene Introns vs ", fraction_name),
    x = expression(log[10]*"(Intron Length)")
  )


# Add Wilcoxon p-value and median differences as annotations
violin_length_plot <- violin_length_plot +
  annotate(
    "text",
    y = 2.2,  # Adjust to position near the test condition, offset
    x = max(log10(plot_data$Length), na.rm = TRUE) * 0.95,
    label = paste0("Wilcoxon p-value: ", format(test_results$wilcox_p_value, digits = 2)),
    size = 3, color = "black", hjust = 0.5
  ) +
  annotate(
    "text",
    y = 1.8,  # Adjust to position near the test condition, offset
    x = max(log10(plot_data$Length), na.rm = TRUE) * 0.95,
    label = paste0("Median length diff: ", format(median_differences$median_diff, digits = 2)),
    size = 3, color = "black", hjust = 0.5
  ) +
  # add n introns for each
 annotate(
    "text",
    y = 2.1,  # Adjust to position near the test condition
    x = 0.5,
    label = paste0("n introns = ", n_test),
    size = 3, color = "black", hjust = 0.5
  ) +
  annotate(
    "text",
    y = 1.1,  # Adjust to position near All_featureCounts
    x = 0.5,
    label = paste0("n introns = ", n_all),
    size = 3, color = "black", hjust = 0.5
  ) 


  print(violin_length_plot)

  #add wd to output_prefix
  output_prefix <- paste0(output_dir, output_prefix) 
  
  # Save the violin plot of intron lengths
  ggsave(paste0(output_prefix, "_Log10_Length_Distribution_", fraction_name, ".png"), plot = violin_length_plot, width = 10, height = 6)
  ggsave(paste0(output_prefix, "_Log10_Length_Distribution_", fraction_name, ".pdf"), plot = violin_length_plot, width = 10, height = 6)


  
}

  
# DESeq2 analysis function for intron data
call_DESeq2_analysis <- function(test_condition, reference_condition, prefix) {
  # Call the DESeq2 analysis function
  res <- run_deseq2_intron_analysis(dds, test_condition, reference_condition, prefix)
  
  # Analyze intron enrichment
  enrichment_data <- analyze_intron_enrichment(res, intron_data, prefix)
  
  # Analyze intron length
  analyze_intron_length(res, intron_data, prefix, test_condition)
  
  return(list(results = res, enrichment_data = enrichment_data))
}


```

```{r DESeq2_Cyto_vs_Cyto GSK591 vs DMSO Introns}
test_condition <- "GSK591_Cytoplasm"
reference_condition <- "DMSO_Cytoplasm"
prefix <- paste0(test_condition, "_vs_", reference_condition)

call_DESeq2_analysis(test_condition, reference_condition, prefix)

```

```{r DESeq2_Nuc_vs_Nuc GSK591 vs DMSO Introns}
test_condition <- "GSK591_Nucleoplasm"
reference_condition <- "DMSO_Nucleoplasm"
prefix <- paste0(test_condition, "_vs_", reference_condition)

call_DESeq2_analysis(test_condition, reference_condition, prefix)

```

```{r DESeq2_Chromatin_vs_Chromatin GSK591 vs DMSO Introns}
test_condition <- "GSK591_Chromatin"
reference_condition <- "DMSO_Chromatin"
prefix <- paste0(test_condition, "_vs_", reference_condition)


call_DESeq2_analysis(test_condition, reference_condition, prefix)


```

```{r DESeq2_DMSO_Chromatin_vs_Cytoplasm  Introns}
test_condition <- "DMSO_Chromatin"
reference_condition <- "DMSO_Cytoplasm"
prefix <- paste0(test_condition, "_vs_", reference_condition)

call_DESeq2_analysis(test_condition, reference_condition, prefix)


```

```{r DESeq2_GSK591_Chromatin_vs_Cytoplasm  Introns}
test_condition <- "GSK591_Chromatin"
reference_condition <- "GSK591_Cytoplasm"
prefix <- paste0(test_condition, "_vs_", reference_condition)


call_DESeq2_analysis(test_condition, reference_condition, prefix)


```


```{r intron_enrichment_analysis, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}

# Function to analyze intron enrichment and plot histogram
analyze_intron_enrichment <- function(res_data, intron_data, output_name, padj_cutoff = 0.05, log2fc_cutoff = 0.58) {
  
  # Filter for significant introns
  sig_introns <- res_data %>%
    filter(padj < padj_cutoff & abs(log2FoldChange) > log2fc_cutoff)
  
  # Extract gene names and count the number of significant introns per gene
  sig_introns <- sig_introns %>%
    mutate(GeneName = sub(":.*", "", IntronID))
  
  sig_intron_counts <- sig_introns %>%
    group_by(GeneName) %>%
    summarize(SigIntronCount = n())
  
  # Extract total intron counts per gene from intron_data
  total_intron_counts <- intron_data %>%
    filter(!grepl("dm6", IntronID)) %>%
    group_by(GeneName) %>%
    summarize(TotalIntronCount = n())
  
  # Merge the significant intron counts with total intron counts
  enrichment_data <- left_join(sig_intron_counts, total_intron_counts, by = "GeneName")
  
  # Calculate the percentage of enriched introns per gene
  enrichment_data <- enrichment_data %>%
    mutate(PercentageEnrichedIntrons = (SigIntronCount / TotalIntronCount) * 100)
  
  # Replace NA values with 0 (genes with no significant introns)
  enrichment_data$PercentageEnrichedIntrons[is.na(enrichment_data$PercentageEnrichedIntrons)] <- 0
  
  # Plot histogram of percentage of enriched introns per gene
  intron_percentage_histogram <- ggplot(enrichment_data, aes(x = PercentageEnrichedIntrons)) +
    geom_histogram(binwidth = 5, fill = "gray", color = "darkgray") +
    theme_minimal() +
    xlim(0, 100) +
    labs(title = paste(output_name, "Percentage of Enriched Introns per Gene"),
         x = "Percentage of Enriched Introns",
         y = "Number of Genes")
  
  print(intron_percentage_histogram)
  
  # Save the histogram
  ggsave(paste0(output_name, "_Percentage_Enriched_Introns_Histogram.png"), plot = intron_percentage_histogram, width = 10, height = 3)
  ggsave(paste0(output_name, "_Percentage_Enriched_Introns_Histogram.pdf"), plot = intron_percentage_histogram, width = 10, height = 3)
  
  return(enrichment_data)
}

# Example usage
#enrichment_data_chromatin <- analyze_intron_enrichment(res_Chromatin, intron_data, "Chromatin")
#enrichment_data_chromatin <- analyze_intron_enrichment(res_Nucleoplasm, intron_data, "Nucleoplasm")
#enrichment_data_chromatin <- analyze_intron_enrichment(res_Cytoplasm, intron_data, "Cytoplasm")
```

```{r intron length analysis and comparison, eval = FALSE}

#function used above
# Function to extract significantly enriched introns for a given fraction, remove dm6 introns
extract_sig_introns <- function(res_data, intron_data, fraction_name) {
  res_data %>%
    filter(padj < 0.05 & log2FoldChange > 0.58) %>%
    dplyr::select(IntronID) %>%
    inner_join(intron_data, by = "IntronID") %>%
    filter(!grepl("dm6", IntronID)) %>%
    mutate(Fraction = fraction_name)
}

# Extract significantly enriched introns for the Chromatin fraction
sig_introns_chromatin <- extract_sig_introns(res_chromatin, intron_data, "Chromatin")

# Create an "all" table from all featureCount genes, remove dm6 introns
all_intron_lengths <- intron_data %>%
  dplyr::select(IntronID, Length) %>%
  filter(!grepl("dm6", IntronID)) %>%
  distinct() %>%
  mutate(Fraction = "All_featureCounts")

# Combine data for "All_featureCounts" and "Chromatin" fractions
plot_data <- bind_rows(
  sig_introns_chromatin %>% dplyr::select(Length, Fraction),
  all_intron_lengths %>% dplyr::select(Length, Fraction)
)

# Set the order of the fractions
plot_data$Fraction <- factor(plot_data$Fraction, levels = c("All_featureCounts", "Chromatin"))

# Calculate the medians for each fraction
medians <- plot_data %>%
  group_by(Fraction) %>%
  summarise(median_length = median(Length))

# Calculate median differences compared to "All_featureCounts"
median_differences <- medians %>%
  filter(Fraction != "All_featureCounts") %>%
  mutate(median_diff = median_length - medians$median_length[medians$Fraction == "All_featureCounts"])

# Perform KS and Wilcoxon tests for the Chromatin fraction against all featureCounts introns
test_results <- plot_data %>%
  filter(Fraction == "Chromatin") %>%
  summarise(
    ks_p_value = ks.test(log10(Length), log10(all_intron_lengths$Length))$p.value,
    wilcox_p_value = wilcox.test(log10(Length), log10(all_intron_lengths$Length))$p.value
  )

# Create a violin plot of the log10 lengths of significantly enriched introns with Wilcoxon p-values and median differences
violin_length_plot <- ggplot(plot_data, aes(x = Fraction, y = log10(Length), fill = Fraction)) +
  geom_violin(trim = FALSE, scale = "width") +
  geom_boxplot(width = 0.1, position = position_dodge(0.9), outlier.shape = NA) +
  theme_minimal() +
  labs(title = "Log10 Length Distribution of Significantly Enriched Introns Compared to All FeatureCounts Introns",
       x = "Fraction",
       y = "Log10(Intron Length)") +
  scale_fill_manual(values = c("All_featureCounts" = "gray", "Chromatin" = "#2D8942")) +
  geom_text(data = test_results, aes(x = "Chromatin", y = max(log10(plot_data$Length)), 
                                     label = paste0("Wilcoxon p-value: ", format(wilcox_p_value, digits = 2))),
            inherit.aes = FALSE, size = 3, color = "black") +
  geom_text(data = median_differences, aes(x = "Chromatin", y = 0.5, 
                                           label = paste0("Median diff: ", format(median_diff, digits = 2))),
            inherit.aes = FALSE, size = 3, color = "black")

print(violin_length_plot)

# Save the violin plot of intron lengths
ggsave("Log10_Length_Distribution_of_Significantly_Enriched_Introns_Compared_to_All_FeatureCounts_Introns_Chromatin.png", plot = violin_length_plot, width = 10, height = 7)
ggsave("Log10_Length_Distribution_of_Significantly_Enriched_Introns_Compared_to_All_FeatureCounts_Introns_Chromatin.pdf", plot = violin_length_plot, width = 10, height = 7)

```

```{r intron-exon per gene}

# Ensure intron and exon data are properly loaded
#intron_data <- process_data(paste0(working_dir,'featureCounts_IntronsExons/Intron_Count_hg38_dm6_A549_frac_PRMTi_s2.txt'), "intron")
#exon_data <- process_data(paste0(working_dir,'featureCounts_IntronsExons/Exon_Count_hg38_dm6_A549_frac_PRMTi_s2.txt'), "exon")

# Aggregate intron and exon counts by gene and condition
aggregate_counts <- function(data, type) {
  long_data <- data %>%
    pivot_longer(cols = starts_with("DMSO") | starts_with("GSK591"), names_to = "Condition_Replicate", values_to = "Count") %>%
    separate(Condition_Replicate, into = c("Condition", "Replicate"), sep = "_rep") %>%
    group_by(GeneName, Condition) %>%
    summarize(avg_count = mean(Count), .groups = 'drop')
  
  long_data <- long_data %>% mutate(Type = type)
  return(long_data)
}

intron_counts <- aggregate_counts(intron_data, "Intron")
exon_counts <- aggregate_counts(exon_data, "Exon")

# Merge intron and exon counts
merged_counts <- inner_join(intron_counts, exon_counts, by = c("GeneName", "Condition"), suffix = c("_intron", "_exon"))

# Calculate intron/exon ratios
merged_counts <- merged_counts %>%
  mutate(IntronExon_Ratio = avg_count_intron / (avg_count_exon + 1))

# Step 1: Aggregate data to ensure unique GeneName and Condition combinations
aggregated_counts <- merged_counts %>%
  group_by(GeneName, Condition) %>%
  summarise(
    IntronExon_Ratio = mean(IntronExon_Ratio, na.rm = TRUE), # Calculate mean ratio per GeneName and Condition
    .groups = "drop"
  )

# Step 2: Pivot the aggregated data to create columns for each Condition
IntronExonRatio_data <- aggregated_counts %>%
  pivot_wider(names_from = Condition, values_from = IntronExon_Ratio)

```

```{r plot comparisons}

# Step 1: Create a function for generating comparison plots
generate_comparison_plot <- function(data, x_column, y_column, 
                                     x_label, y_label, 
                                     title, output_filename,
                                     x_limits = NULL, y_limits = NULL) {
  comparison_plot <- ggplot(data, aes_string(x = x_column, y = y_column)) +
    geom_pointdensity(size = 0.5) +
    scale_color_viridis() +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12, colour = "black"),
      axis.text.x = element_text(size = 12, colour = "black"),
      axis.title.y = element_text(size = 12, colour = "black"),
      axis.title.x = element_text(size = 12, colour = "black"),
      axis.ticks = element_line(colour = "black"),
      axis.ticks.length = unit(0.1, "cm"),
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      legend.position = "right",
      aspect.ratio = 1
    ) +
    labs(
      title = title,
      x = x_label,
      y = y_label,
      color = "Transcript Density"
    )
    # Apply x and y limits if provided
  if (!is.null(x_limits)) {
    comparison_plot <- comparison_plot + scale_x_continuous(limits = x_limits)
  }
  if (!is.null(y_limits)) {
    comparison_plot <- comparison_plot + scale_y_continuous(limits = y_limits)
  }
  # Display the plot
  print(comparison_plot)

  # Save the plot
  ggsave(filename = paste0(output_dir, output_filename, ".png"), 
         plot = comparison_plot, width = 8, height = 8)
  ggsave(filename = paste0(output_dir, output_filename, ".pdf"), 
         plot = comparison_plot, width = 8, height = 8)
}

# Step 2: Add natural log-transformed columns for all conditions
IntronExonRatio_data <- IntronExonRatio_data %>%
  mutate(
    ln_DMSO_Chromatin = log(DMSO_Chromatin + 1),
    ln_DMSO_Cytoplasm = log(DMSO_Cytoplasm + 1),
    ln_DMSO_Nucleoplasm = log(DMSO_Nucleoplasm + 1),
    ln_GSK591_Chromatin = log(GSK591_Chromatin + 1),
    ln_GSK591_Cytoplasm = log(GSK591_Cytoplasm + 1),
    ln_GSK591_Nucleoplasm = log(GSK591_Nucleoplasm + 1)
  )
# Step 3: Call the function for specific comparisons
# Comparison 1: DMSO Chromatin vs GSK591 Chromatin
generate_comparison_plot(
  data = IntronExonRatio_data,
  x_column = "ln_DMSO_Chromatin",
  y_column = "ln_GSK591_Chromatin",
  x_label = "ln(DMSO Chromatin Intron/Exon Ratio)",
  y_label = "ln(GSK591 Chromatin Intron/Exon Ratio)",
  title = "IntronExonRatio: DMSO Chromatin vs GSK591 Chromatin (per transcript)",
  output_filename = "IntronExonRatio_DMSO-Chromatin_vs_GSK591_Chromatin",
  x_limits = c(0,4),
  y_limits = c(0,4)
)

# Comparison 2: DMSO Chromatin vs DMSO Cytoplasm
generate_comparison_plot(
  data = IntronExonRatio_data,
  x_column = "ln_DMSO_Chromatin",
  y_column = "ln_DMSO_Cytoplasm",
  x_label = "ln(DMSO Chromatin Intron/Exon Ratio)",
  y_label = "ln(DMSO Cytoplasm Intron/Exon Ratio)",
  title = "IntronExonRatio: DMSO Chromatin vs DMSO Cytoplasm (per transcript)",
  output_filename = "IntronExonRatio_DMSO_Chromatin_vs_DMSO_Cytoplasm",
    x_limits = c(0,4),
  y_limits = c(0,4)
)

# Comparison 2: DMSO Chromatin vs DMSO Cytoplasm
generate_comparison_plot(
  data = IntronExonRatio_data,
  x_column = "ln_DMSO_Chromatin",
  y_column = "ln_DMSO_Nucleoplasm",
  x_label = "ln(DMSO Chromatin Intron/Exon Ratio)",
  y_label = "ln(DMSO Nucleoplasm Intron/Exon Ratio)",
  title = "IntronExonRatio: DMSO Chromatin vs DMSO Nucleoplasm (per transcript)",
  output_filename = "IntronExonRatio_DMSO_Chromatin_vs_DMSO_Nucleoplasm",
    x_limits = c(0,4),
  y_limits = c(0,4)
)

# Comparison 3: GSK591 Chromatin vs GSK591 Cytoplasm
generate_comparison_plot(
  data = IntronExonRatio_data,
  x_column = "ln_DMSO_Cytoplasm",
  y_column = "ln_GSK591_Cytoplasm",
  x_label = "ln(GSK591 Cytoplasm Intron/Exon Ratio)",
  y_label = "ln(GSK591 Cytoplasm Intron/Exon Ratio)",
  title = "IntronExonRatio: DMSO Cytoplasm vs GSK591 Cytoplasm  (per transcript)",
  output_filename = "IntronExonRatio_DMSO_Cytoplasm_vs_GSK591_Cytoplasm",
    x_limits = c(0,4),
  y_limits = c(0,4)
)


```

```{r IntronExon Ratio Comparison Plots}

# Step 1: Read GRIPPs file and prepare for overlay
GRIPPs <- read.csv("C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 7/GRIPPs/GRIPPs.csv")

# Filter IntronExonRatio_data for genes in GRIPPs
IntronExonRatio_GRIPPs <- IntronExonRatio_data %>%
  filter(GeneName %in% gripps$Symbol) %>%
  mutate(is_GRIPP = TRUE)

# Combine GRIPPs data with the full dataset
IntronExonRatio_full <- IntronExonRatio_data %>%
  mutate(is_GRIPP = ifelse(GeneName %in% gripps$Symbol, TRUE, FALSE))

# Step 2: Updated function for overlaying GRIPPs
generate_comparison_plot_with_overlay <- function(data, overlay_data, 
                                                  x_column, y_column, 
                                                  x_label, y_label, 
                                                  title, output_filename,
                                                   x_limits = NULL, y_limits = NULL) {
  comparison_plot <- ggplot(data, aes_string(x = x_column, y = y_column)) +
    # Base data in gray
    geom_point(color = "#D4D4D4", alpha = 0.5, size = 0.5) +
    # Overlay GRIPPs in transparent purple
    geom_point(data = overlay_data, aes_string(x = x_column, y = y_column), 
               color = "#666666", alpha = 0.5, size = 1) +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12, colour = "black"),
      axis.text.x = element_text(size = 12, colour = "black"),
      axis.title.y = element_text(size = 12, colour = "black"),
      axis.title.x = element_text(size = 12, colour = "black"),
      axis.ticks = element_line(colour = "black"),
      axis.ticks.length = unit(0.1, "cm"),
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      legend.position = "none",
      aspect.ratio = 1
    ) +
    labs(
      title = title,
      x = x_label,
      y = y_label
    )
    # Apply x and y limits if provided
  if (!is.null(x_limits)) {
    comparison_plot <- comparison_plot + scale_x_continuous(limits = x_limits)
  }
  if (!is.null(y_limits)) {
    comparison_plot <- comparison_plot + scale_y_continuous(limits = y_limits)
  }
  # Display the plot
  print(comparison_plot)
  # Display the plot
  print(comparison_plot)
  
  # Save the plot
  ggsave(filename = paste0(output_dir, output_filename, ".png"), 
         plot = comparison_plot, width = 8, height = 8)
  ggsave(filename = paste0(output_dir, output_filename, ".pdf"), 
         plot = comparison_plot, width = 8, height = 8)
}

# Step 3: Call the function for the desired comparison
generate_comparison_plot_with_overlay(
  data = IntronExonRatio_data,
  overlay_data = IntronExonRatio_GRIPPs,
  x_column = "ln_DMSO_Chromatin",
  y_column = "ln_DMSO_Cytoplasm",
  x_label = "ln(DMSO Chromatin Intron/Exon Ratio)",
  y_label = "ln(DMSO Cytoplasm Intron/Exon Ratio)",
  title = "IntronExonRatio: DMSO Chromatin vs DMSO Cytoplasm (with GRIPPs overlay)",
  output_filename = "IntronExonRatio_DMSO_Chromatin_vs_DMSO_Cytoplasm_with_GRIPPs",
  x_limits = c(0,4),
  y_limits = c(0,4)
)


```

```{r PRMT5 dependent GRIPPs}
# Step 1: Load DESeq2 results and filter for GSK591-dependent GRIPPs
PRMT5GRIPPs_results <- read.csv("C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 6/Output/GSK591_chromatin_vs_DMSO_chromatin_SpikeInNormalized_DESEQ2.csv")

# Filter for significant GSK591-dependent GRIPPs
prmt5_dependent_gripps <- PRMT5GRIPPs_results %>%
  dplyr::filter(padj < 0.01 & log2FoldChange > 1) %>%
  dplyr::select(GeneSymbol) %>%
  dplyr::rename(Symbol = GeneSymbol) %>% # Rename GeneSymbol to Symbol
  distinct()

# GRIPPs
IntronExonRatio_GRIPPs <- IntronExonRatio_data %>%
  filter(GeneName %in% gripps$Symbol) %>%
  mutate(is_GRIPP = TRUE)

# PRMT5-dependent GRIPPs
IntronExonRatio_PRMT5_GRIPPs <- IntronExonRatio_data %>%
  filter(GeneName %in% prmt5_dependent_gripps$Symbol) %>%
  mutate(is_PRMT5_GRIPP = TRUE)

# Step 3: Combine GRIPPs and PRMT5-dependent GRIPPs with the full dataset
IntronExonRatio_full <- IntronExonRatio_data %>%
  mutate(
    is_GRIPP = ifelse(GeneName %in% gripps$Symbol, TRUE, FALSE),
    is_PRMT5_GRIPP = ifelse(GeneName %in% prmt5_dependent_gripps$Symbol, TRUE, FALSE)
  )

# Step 4: Define a new function for multiple overlays
generate_comparison_plot_with_multiple_overlays <- function(data, overlay_data1, overlay_data2, 
                                                            x_column, y_column, x_label, y_label, 
                                                            title, output_filename, 
                                                            x_limits = NULL, y_limits = NULL) {
  comparison_plot <- ggplot(data, aes_string(x = x_column, y = y_column)) +
    # Base data in gray
    geom_point(color = "#DADADA", alpha = 0.5, size = 1) +
    # Overlay GRIPPs in purple
    geom_point(data = overlay_data1, aes_string(x = x_column, y = y_column), 
               color = "#666666", alpha = 0.7, size = 1) +
    # Overlay PRMT5-dependent GRIPPs in blue
    geom_point(data = overlay_data2, aes_string(x = x_column, y = y_column), 
               color = "#0D8943", alpha = 0.7, size = 1) +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12, colour = "black"),
      axis.text.x = element_text(size = 12, colour = "black"),
      axis.title.y = element_text(size = 12, colour = "black"),
      axis.title.x = element_text(size = 12, colour = "black"),
      axis.ticks = element_line(colour = "black"),
      axis.ticks.length = unit(0.1, "cm"),
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      legend.position = "none",
      aspect.ratio = 1
    ) +
    labs(
      title = title,
      x = x_label,
      y = y_label
    )
   # Apply x and y limits if provided
  if (!is.null(x_limits)) {
    comparison_plot <- comparison_plot + scale_x_continuous(limits = x_limits)
  }
  if (!is.null(y_limits)) {
    comparison_plot <- comparison_plot + scale_y_continuous(limits = y_limits)
  }
  # Display the plot
  print(comparison_plot)
  
  # Save the plot
  ggsave(filename = paste0(output_dir, output_filename, ".png"), 
         plot = comparison_plot, width = 8, height = 8)
  ggsave(filename = paste0(output_dir, output_filename, ".pdf"), 
         plot = comparison_plot, width = 8, height = 8)
}

# Step 5: Call the function for the desired comparison
generate_comparison_plot_with_multiple_overlays(
  data = IntronExonRatio_data,
  overlay_data1 = IntronExonRatio_GRIPPs,
  overlay_data2 = IntronExonRatio_PRMT5_GRIPPs,
  x_column = "ln_DMSO_Chromatin",
  y_column = "ln_DMSO_Cytoplasm",
  x_label = "ln(DMSO Chromatin Intron/Exon Ratio)",
  y_label = "ln(DMSO Cytoplasm Intron/Exon Ratio)",
  title = "IntronExonRatio: DMSO Chromatin vs DMSO Cytoplasm (with GRIPPs and PRMT5-dependent GRIPPs overlays)",
  output_filename = "IntronExonRatio_DMSO_Chromatin_vs_DMSO_Cytoplasm_with_GRIPPs_and_PRMT5",
  x_limits = c(0,3),
  y_limits = c(0,3)
)

```