---
title: "Define and Analyze GRIPPs"
author: "David Shechter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: default
  pdf_document: default
---
# GRIPPs Definition and Analysis

## Overview

## Criteria for Defining GRIPPs

This script defines and analyzes **Genomically-Retained Incompletely Processed Polyadenylated Transcripts (GRIPPs)** using RNA-seq data from A549 cells fractionated into chromatin and cytoplasmic compartments under DMSO control conditions. GRIPPs are identified as transcripts that exhibit a significant enrichment in the chromatin fraction compared to the cytoplasm, normalized using spike-in controls.



```{r setup, include=FALSE}
# Load necessary libraries
library(tidyverse)
library(biomaRt)
library(readr)
library(ggplot2)
library(dplyr)

# Set working directory
# Set the WD here
working_dir <- 'C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 7/'
knitr::opts_knit$set(root.dir = normalizePath(working_dir))
setwd(working_dir)
getwd()

# Create output directory
output_dir <- paste0(working_dir, "GRIPPs/")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir)
  }

# Input and output file paths
input_file <- "C:/Users/david/OneDrive/Documents/GitHub/DeAngelo_2024/Figure 6/Output/DMSO_chromatin_vs_DMSO_cytoplasm_SpikeInNormalized_DESEQ2.csv"
gripps_csv <- "./GRIPPs/GRIPPs.csv"
gripps_bed <- "./GRIPPs/GRIPPs.bed"
```

### Define GRIPPs Criteria

```{r define-gripps}
# Read the DESeq2 results file
deseq_data <- read_csv(input_file)

# Define GRIPPs: high chromatin enrichment and statistically significant
# Adjust these thresholds as necessary
log2fc_cutoff <- 0.58  # Log2 fold-change threshold
pvalue_cutoff <- 0.01  # Adjusted p-value threshold

gripps <- deseq_data %>%
  filter(log2FoldChange > log2fc_cutoff & padj < pvalue_cutoff) %>%
  dplyr::select(ENSG = gene_id, Symbol = GeneSymbol) %>%
  distinct()

# Write GRIPPs.csv file
write_csv(gripps, gripps_csv)
```

### Retrieve Gene Coordinates

```{r gene-coordinates, include = FALSE}

# Use biomaRt to fetch gene coordinates from Ensembl
mart <- useMart(
  biomart = "ENSEMBL_MART_ENSEMBL",
  dataset = "hsapiens_gene_ensembl",
  host = "https://www.ensembl.org"
)

# Get gene coordinates
coordinates <- getBM(
  attributes = c(
    "ensembl_gene_id", "hgnc_symbol", "chromosome_name",
    "start_position", "end_position", "strand", "gene_biotype"
  ),
  filters = "ensembl_gene_id",
  values = gripps$ENSG,
  mart = mart
)

# Rename and format coordinates
coordinates <- coordinates %>%
  dplyr::rename(
    Chromosome = chromosome_name,
    Start = start_position,
    End = end_position,
    Strand = strand,
    Biotype = gene_biotype
  ) %>%
  dplyr::mutate(
    Score = 0,  # Add a placeholder score column for BED format
    Strand = ifelse(Strand == 1, "+", "-")  # Convert strand to BED format
  )

# Create GRIPPs.csv file
gripps <- coordinates %>%
  dplyr::select(ensembl_gene_id = ensembl_gene_id, Symbol = hgnc_symbol, Biotype)

write.csv(
  gripps,
  file = "GRIPPs.csv",
  quote = FALSE,
  row.names = FALSE
)

#Create bed file for GRIPPs
GRIPPs_bed <- coordinates %>%
  dplyr::select(Chromosome, Start, End, ENSG = ensembl_gene_id, Symbol = hgnc_symbol, Strand)

write.table(
  GRIPPs_bed,
  file = "GRIPPs.bed",
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

```

### Summary

```{r summary}
# Display a summary of the GRIPPs
cat("Number of GRIPPs identified:", nrow(gripps), "\n")
head(gripps)

# Combine pseudogenes and TEC into distinct categories
gene_type_summary <- gripps %>%
  mutate(
    RNA_Class = case_when(
      grepl("pseudogene", Biotype, ignore.case = TRUE) ~ "Pseudogene",
      Biotype == "TEC" ~ "TEC",
      TRUE ~ Biotype  # Keep the original Biotype for others
    )
  ) %>%
  group_by(RNA_Class) %>%
  summarize(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

# Generate the stacked bar plot
gene_class_plot <- ggplot(gene_type_summary, aes(x = "", y = Percentage, fill = RNA_Class)) +
  geom_bar(stat = "identity", width = 0.5, color = "black") +
  coord_flip() +
  theme_minimal() +
  scale_fill_manual(values = c(
          "protein_coding" = "#a6cee3",
      "lncRNA" = "#1f78b4",
      "Pseudogene" = "#b2df8a",
      "TEC" = "#33A02C",
      "other" = "#999999")) +
  labs(
    title = "Transcript Classes in GRIPPs",
    x = NULL,
    y = "Percentage of RNA Classes",
    fill = "RNA Class"
  ) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

# Print the plot
print(gene_class_plot)


```

```{r GRIPPs clusterProfiler, eval = FALSE}
# Load libraries
library(clusterProfiler)
library(org.Hs.eg.db)  # Human genome annotations
library(enrichplot)

# Convert GRIPPs identifiers to ENTREZID
GRIPPs_ENTREZ <- bitr(
  gripps$ensembl_gene_id, 
  fromType = "ENSEMBL", 
  toType = "ENTREZID", 
  OrgDb = org.Hs.eg.db
)

# Create the universe: All genes with valid ENTREZID in your dataset
universe_genes <- unique(bitr(
  deseq_data$gene_id, 
  fromType = "ENSEMBL", 
  toType = "ENTREZID", 
  OrgDb = org.Hs.eg.db
)$ENTREZID)

# Ensure GRIPPs_ENTREZ is a subset of the universe
GRIPPs_ENTREZ <- GRIPPs_ENTREZ$ENTREZID[GRIPPs_ENTREZ$ENTREZID %in% universe_genes]

# GO enrichment analysis (Biological Process)
go_enrichment_BP <- enrichGO(
  gene = GRIPPs_ENTREZ, 
  OrgDb = org.Hs.eg.db, 
  ont = "BP",  # Biological Process
  universe = universe_genes, 
  pAdjustMethod = "BH", 
  pvalueCutoff = 0.05, 
  qvalueCutoff = 0.05, 
  keyType = "ENTREZID"
)

# GO enrichment analysis (Molecular Function)
go_enrichment_MF <- enrichGO(
  gene = GRIPPs_ENTREZ, 
  OrgDb = org.Hs.eg.db, 
  ont = "MF",  
  universe = universe_genes, 
  pAdjustMethod = "BH", 
  pvalueCutoff = 0.05, 
  qvalueCutoff = 0.05, 
  keyType = "ENTREZID"
)

# KEGG pathway enrichment analysis (optional)
kegg_enrichment <- enrichKEGG(
  gene = GRIPPs_ENTREZ, 
  organism = 'hsa', 
  universe = universe_genes, 
  pAdjustMethod = "BH", 
  pvalueCutoff = 0.05
)

# Plot results
if (!is.null(go_enrichment_BP) && nrow(go_enrichment) > 0) {
  print(dotplot(go_enrichment, showCategory = 5) + ggtitle("GO BP Enrichment: GRIPPs"))
} else {
  message("No significant GO BP terms found.")
}

if (!is.null(go_enrichment_MF) && nrow(go_enrichment) > 0) {
print(dotplot(go_enrichment_MF, showCategory = 5) + ggtitle("GO MF Enrichment: GRIPPs"))
} else {
  message("No significant GO MF terms found.")
}
 
if (!is.null(kegg_enrichment) && nrow(kegg_enrichment) > 0) {
  print(dotplot(kegg_enrichment, showCategory = 5) + ggtitle("KEGG Enrichment: GRIPPs"))
} else {
  message("No significant KEGG pathways found.")
}



# Save results
write.csv(as.data.frame(go_enrichment), "GRIPPs_GO_Enrichment.csv", row.names = FALSE)
write.csv(as.data.frame(kegg_enrichment), "GRIPPs_KEGG_Enrichment.csv", row.names = FALSE)


```

### Output Files
- `GRIPPs.csv`: Contains the ENSG and Symbol of GRIPPs.
- `GRIPPs.bed`: BED file with complete gene coordinates in hg38.

Run this script to generate the outputs in the specified working directory.
